---
title: "Nutrient Data Integration Code"
author: "Morgan Battey"
date: "2025-03-13"
output: html_document
---

## Load Packages

Load necessary packages that we will use for this analysis

```{r}
library(tidyverse)
library(glue)
library(EDIutils)
library(here)
```

## Run Functions

Run function to grab EDI data

```{r}
get_edi_file <- function(pkg_id, fname) {
  # define scope
  scope = 'edi'
  
  # get latest revision
  revisions <- EDIutils::list_data_package_revisions(scope = scope, identifier = pkg_id)
  latest_revision <- max(as.numeric(revisions))
  package_id_str <- glue::glue('{scope}.{pkg_id}.{latest_revision}')

  # get entity IDs
  entities <- EDIutils::list_data_entities(packageId = package_id_str)

  # slow wrapper (avoid rate limit)
  slow_read <- purrr::slowly(EDIutils::read_data_entity_name, purrr::rate_delay(pause = 1))

  # find the matching entity
  matched <- purrr::keep(entities, function(entity_id) {
    entity_name <- slow_read(packageId = package_id_str, entityId = entity_id)
    identical(entity_name, fname)
  })

  if (length(matched) == 0) {
    stop(glue::glue("File '{fname}' not found in package edi.{pkg_id}.{latest_revision}"))
  }

  # construct download URL and read csv
  entity_id <- matched[[1]]
  file_url <- glue::glue('https://pasta.lternet.edu/package/data/eml/{scope}/{pkg_id}/{latest_revision}/{entity_id}')
  df <- readr::read_csv(file_url, guess_max = 1000000, show_col_types = FALSE)

  return(df)
}
```

Run function to read in SharePoint files

```{r}
abs_nutrient_path <- function(fp_rel = NULL) {
  base_path <- file.path(
    Sys.getenv('USERPROFILE'), 
    'California Department of Water Resources', 
    'Nutrient synthesis - Documents/Data Integration Team/2025 Data Integration Code Files'
  )

  if (is.null(fp_rel)) {
    return(base_path)
  } else {
    return(file.path(base_path, fp_rel))
  }
}
```

Run function to check for coerced NAs

```{r}
check_non_numeric <- function(df, col_name) {
  if (col_name %in% names(df)) {
    original <- df[[col_name]]
    converted <- as.numeric(as.character(original))
    # Find which values became NA
    na_indices <- which(is.na(converted) & !is.na(original))
    if (length(na_indices) > 0) {
      cat('Column:', col_name, '\n')
      cat('Problematic values:\n')
      print(unique(original[na_indices]))
      cat('At rows:', na_indices[1:min(10, length(na_indices))], '\n\n')
    } else {
    cat('Column:', col_name, 'has no problematic values\n')
    }
  } 
}
```

## WDL Data

### 1 - Tidal Parr

Import Tidal Parr data

```{r}
fp_tp <- abs_nutrient_path("Tidal Parr_Data_Final.csv")

df_tp <- read_csv(fp_tp) #read in data file

rm(fp_tp) #remove file paths
```

Check/Format Data

```{r}
unique(df_tp$Matrix) #only shows Water, Natural, but also has NA's... why? 
#df_tp %>% filter(is.na(Matrix)) %>% View()
#looks like a few blanks and random lines with stuff we don't want. 
#filtering to remove rows with NA's for results (below) should fix this. 

unique(df_tp$`Sample Type`) #only shows Normal Samples
unique(df_tp$`Lab Dup`) #but there are still lab duplicates in there, which is confusing. Maybe filter those out?
unique(df_tp$`Analyte`) #no other analytes present besides what we want 

#Morgan's response to Lab Dup question: Sample Type column only indicates normal, field dups, and blanks (does not indicate lab dups), everything except normal samples gets filtered out later, including lab dups 

df_tp_final <- df_tp %>%
  mutate(Date_Time = mdy_hm(`Collection Date`)) %>% #format date
  select(c("Station Number", "Date_Time", "Sample Code", "Analyte", "Result", "Rpt Limit", "Units", "Method", "Lab Dup", "Sample Type", "Description", "Notes")) %>% #remove unwanted columns
  add_column(Project = "Tidal Parr Study") %>% #add Project column and have it all say Tidal Parr
  add_column("Long Station Name" = NA, "Short Station Name" = NA) %>% #data file doesn't have station names
  filter(!(is.na(`Result`) | `Result` == "N.S.")) %>% #get rid of rows with NA or NS in the result column
  filter(is.na(`Lab Dup`)) # remove lab duplicates

# double check unique checks now that NA results are gone and lab dupes are filtered out
unique(df_tp_final$`Sample Type`)
unique(df_tp_final$`Lab Dup`)
unique(df_tp_final$`Analyte`)

rm(df_tp) #remove intermediate data frames
```

Check DateTimes

```{r}
# any missing dates?
#df_tp_final %>% filter(is.na(Date_Time)) %>% View()

# no missing dates

# any missing times?
#df_tp_final %>% filter(format(Date_Time, '%H:%M:%S') == '00:00:00') %>% View()

#no missing date times
```

### 2 - O&M

Import O&M data (multiple files)

```{r}
fp_aq1 <- abs_nutrient_path("O&M Data_CA Aquaduct Grab Sample_1.csv")
df_aq1 <- read_csv(fp_aq1)

fp_aq2 <- abs_nutrient_path("O&M Data_CA Aquaduct Grab Sample_2.csv")
df_aq2 <- read_csv(fp_aq2)

fp_aq3 <- abs_nutrient_path("O&M Data_CA Aquaduct Grab Sample_3.csv")
df_aq3 <- read_csv(fp_aq3)

fp_nba <- abs_nutrient_path("O&M Data_NBA Grab Sample.csv")
df_nba <- read_csv(fp_nba)

fp_ofd <- abs_nutrient_path("O&M Data_OFD Grab Sample.csv")
df_ofd <- read_csv(fp_ofd)

fp_sba <- abs_nutrient_path("O&M Data_SBA Grab Sample.csv")
df_sba <- read_csv(fp_sba)

rm(fp_aq1, fp_aq2, fp_aq3, fp_nba, fp_ofd, fp_sba) #remove file paths
```

Combine data files

```{r}
df_om <- bind_rows(df_aq1, df_aq2, df_aq3, df_nba, df_ofd, df_sba) #combine all data files into one data frame

# check nothing was lost in the bind
nrow(df_om) == nrow(df_aq1)+nrow(df_aq2)+nrow(df_aq3)+nrow(df_nba)+nrow(df_ofd)+nrow(df_sba)

rm(df_aq1, df_aq2, df_aq3, df_nba, df_ofd, df_sba) #remove intermediate data frames
#Just a random note from Rosie, but I almost never remove intermediate data frames unless I start running into memory issues. But maybe it's best practice?
# Perry: I like to for clutter reasons (after ensuring everything works correctly), but yeah, it's personal pref so long as you're not running into memory problems.
```

Check/Format Data

```{r}
unique(df_om$Matrix) #only shows Water, Natural
unique(df_om$`Sample Type`) #only shows Normal Samples
unique(df_om$Analyte) #no other analytes present besides what we want
#df_om %>% filter(is.na(Matrix)) %>% View()
#just junk columns left over at the end.

unique(df_om$`Lab Dup`) #still some lab dups in there

df_om_new <- df_om %>% filter(!(str_detect(`Sample Code`, "^C"))) #remove the sample codes that start with a "C" (these belong to a different group and were not sampled by O&M - confirmed with data contact)
  
df_om_final <- df_om_new %>%
  mutate(Date_Time = mdy_hm(`Collection Date`)) %>% #format date
  select(c("Long Station Name", "Short Station Name", "Station Number", "Date_Time", "Sample Code", "Analyte", "Result", "Rpt Limit", "Units", "Method", "Lab Dup", "Sample Type", "Description", "Notes")) %>% #remove unwanted columns
  add_column(Project = "SWP Routine Monitoring") %>% #add Project column and have it all say SWP Routine Monitoring
  filter(!(is.na(`Result`) | `Result` == "N.S.")) %>% #get rid of rows with NA or NS in the result column
  filter(is.na(`Lab Dup`)) # remove lab duplicates

# double check after filtering
unique(df_om_final$`Sample Type`)
unique(df_om_final$Analyte) 
unique(df_om_final$`Lab Dup`)


rm(df_om, df_om_new) #remove intermediate data frames
```

Check DateTimes

```{r}
# any missing dates?
#df_om_final %>% filter(is.na(Date_Time)) %>% View()

# any missing times?
#df_om_final %>%
  #filter(format(Date_Time, '%H:%M:%S') == '00:00:00') %>%
  #distinct(`Date_Time`,`Sample Code`) %>% View()

# a few missing times
# TODO: follow up? or is it fine?
#Morgan: I think it's probably fine for discrete
```

### 3 - NDFS (DISE)

Import NDFS data

```{r}
#looks like the first 1000 "Notes" fields were blank, so got read in as logical.
fp_ndfs <- abs_nutrient_path("NDFS Data.csv")
df_ndfs <- read_csv(fp_ndfs, guess_max = 4000)

rm(fp_ndfs) #remove file paths
```

Check/Format data

```{r}
unique(df_ndfs$Matrix) #only shows Water, Natural
unique(df_ndfs$`Sample Type`) #shows Normal Samples and Duplicate Samples
unique(df_ndfs$`Analyte`) #no other analytes present besides what we want

df_dise_ndfs_final <- df_ndfs %>%
  mutate(Date_Time = mdy_hm(`Collection Date`)) %>% #format date
  select(c("Long Station Name", "Short Station Name", "Station Number", "Date_Time", "Sample Code", "Analyte", "Result", "Rpt Limit", "Units", "Method", "Lab Dup", "Sample Type", "Description", "Notes")) %>% #remove unwanted rows
  add_column(Project = "DISE NDFS Study") %>% #add Project column and have it all say NDFS Study
  filter(!(is.na(`Result`) | `Result` == "N.S.")) %>% #get rid of rows with NA or NS in the result column
  filter(`Sample Type` == 'Normal Sample') # filter out duplicates

rm(df_ndfs) #remove intermediate data frames
```

Update Liberty Island station

```{r}
#df_dise_ndfs_final %>% distinct(`Short Station Name`, `Station Number`) %>% View()

#DISE NDFS and WQA have the same coordinates but slightly different names for South Tip at Liberty Island - use WQA's station name and number (confirmed this is okay with data contacts)

df_dise_ndfs_final <- df_dise_ndfs_final %>% 
  # change to the WQA station number
  mutate(
    "Station Number" = if_else(
      `Station Number` == "B9D81450411", 
      "B9D81461410", `Station Number`
    )
  )
```

Check DateTimes

```{r}
# any missing dates?
#df_dise_ndfs_final %>% filter(is.na(Date_Time)) %>% View()

# any missing times?
#df_dise_ndfs_final %>%
  #filter(format(Date_Time, '%H:%M:%S') == '00:00:00') %>%
  #distinct(`Date_Time`,`Sample Code`) %>% View()
```

### 4 - NCRO

Import NCRO data (all projects combined into one file)

```{r}
fp_ncro <- abs_nutrient_path("NCRO Data.csv")
df_ncro <- read_csv(fp_ncro, guess_max = 30000)

rm(fp_ncro) #remove file paths
```

Add project names

```{r}
unique(df_ncro$`Short Station Name`)
unique(df_ncro$`Long Station Name`)

# only NA is junk data; filter out
df_ncro %>% filter(is.na(`Long Station Name`) | is.na(`Short Station Name`)) %>% View()

df_ncro <- df_ncro %>% filter(!(is.na(`Long Station Name`) | is.na(`Short Station Name`))) # two fewer, as expected

#since all projects were exported from WDL in one file, we need to differentiate them based on station names

#project 1 - south delta temporary barriers
df_sdtb <- df_ncro %>%
  filter(`Long Station Name` == "Doughty Cut near Grantline Canal" | `Long Station Name` == "Grant Line Canal near Old River" | `Long Station Name` == "Grant Line Canal East" | `Long Station Name` == "Middle River at Howard Road" | `Long Station Name` == "Middle River at Undine Road" | `Long Station Name` == "Middle River near Tracy Road" | `Long Station Name` == "Old River above DMC Barrier" | `Long Station Name` == "Old River near Head" | `Long Station Name` == "Old River Upstream of Mountain House Creek" | `Long Station Name` == "Old River near Doughty Cut - ORX" | `Long Station Name` == "Paradise Cut Upstream" | `Long Station Name` == "Sugar Cut downstream of Tom Paine Slough" | `Long Station Name` == "Old River at Tracy Wildlife Association" | `Long Station Name` == "Victoria Canal near Byron" | `Long Station Name` == "Tom Paine Slough above Intake" | `Long Station Name` == "Grant Ln Can @ Tracy Rd Bdg" | `Long Station Name` == "Grantline Canal above Dam" | `Long Station Name` == "Middle R @ Tracy Rd Bdg" | `Long Station Name` == "Middle River @ Union Point - P10A" | `Long Station Name` == "Old River Downstream DMC Barrier") %>% #filter for the SDTB stations
  add_column(Project = "South Delta Temporary Barriers Project") #add Project column and have it all say SDTB

#project 2 - emergency drought barrier 
df_edb <- df_ncro %>%
  filter(`Short Station Name` == "Bethel near Piper" | `Short Station Name` == "False R Nr Oakley" | `Short Station Name` == "Fisherman's Cut" | `Short Station Name` == "Middle River Nr Holt" | `Short Station Name` == "Holland Cut Nr BI" | `Short Station Name` == "Miner Slough nr SR" | `Short Station Name` == "OR Nr Franks Tract" | `Short Station Name` == "Sac R. DS Isleton" | `Short Station Name` == "Steamboat Slough" | `Short Station Name` == "Three Mile Slough") %>% #filter for the EDB stations
  add_column(Project = "Emergency Drought Barrier") #add Project column and have it all say EDB

#project 3 - rock slough WQ monitoring program
df_rs <- df_ncro %>%
  filter(`Short Station Name` == "RS@DRB" | `Short Station Name` == "RS @CCWD Fish Screen") %>% #filter for the rock slough stations
  add_column(Project = "Rock Slough Water Quality Monitoring Program") #add Project column and have it all say rock slough

#project 4 - north delta food subsidies
df_ndfs <- df_ncro %>%
  filter(`Short Station Name` == "Toe Drain YB LISBON") %>%
  add_column(Project = "NCRO NDFS Study") #add Project column and have it all say NCRO NDFS

#combine the data frames into one
df_ncro_new <- bind_rows(df_sdtb, df_edb, df_rs, df_ndfs)
```

Double check all stations were assigned correctly

```{r}
# differing number of rows; ncro_new missing about 30 rows - fixed!
nrow(df_ncro_new) == nrow(df_ncro)

# same 3 stations missing
setdiff(df_ncro$`Long Station Name`, df_ncro_new$`Long Station Name`)
setdiff(df_ncro$`Short Station Name`, df_ncro_new$`Short Station Name`)

# TODO: should these be included in the dataset?
#Morgan: Yes, I missed those. Added them in above to the sdtb data frame

# Perry tip: usually better to use case_when in mutate than to break up in multiple dataframes, as it's easier to tell when data is excluded/formatted incorrectly. Can you "%in%" so you don't have to write out the column name each time, eg:
# df_test %>% mutate(Column = case_when(
                # Other_col %in% c(<list of stations>) ~ <val to assign>,
                # TRUE ~ NA)
              # )

# Dave's additional tip: another method to streamline assigning projects is to create a lookup table
# of station names and their assigned projects, then left_join to df_ncro
```

Check/format data

```{r}
unique(df_ncro_new$Matrix) #only shows Water, Natural
unique(df_ncro_new$`Sample Type`) #shows Normal Samples, Replicate Samples, and Duplicate Samples
unique(df_ncro_new$`Analyte`) #includes the analytes we want, plus dissolved nitrate (keep) and dissolved TKN (remove), and includes field data


df_ncro_final <- df_ncro_new %>%
  mutate(Date_Time = mdy_hm(`Collection Date`)) #format date

range(df_ncro_final$Date_Time) #dates are from 1999-2024, need to remove everything before 2010

df_ncro_final <- df_ncro_final %>% 
  mutate(year = year(Date_Time)) %>% #create a year column
  filter(year > 2009) %>% #remove everything before 2010
  filter(Analyte == "Dissolved Ammonia" | Analyte == "Dissolved Nitrate + Nitrite"| Analyte == "Dissolved Organic Nitrogen"| Analyte == "Dissolved ortho-Phosphate" | Analyte == "Dissolved Nitrate" | Analyte == "Total Phosphorus" | Analyte == "Total Kjeldahl Nitrogen") %>% #filter for the analytes we want
  select(c("Long Station Name", "Short Station Name", "Station Number", "Date_Time", "Sample Code", "Analyte", "Result", "Rpt Limit", "Units", "Method", "Lab Dup", "Sample Type", "Description", "Notes", Project)) %>% #remove unwanted rows
  filter(!(is.na(`Result`) | `Result` == "N.S.")) #get rid of rows with NA or NS in the result column

# check this filtered correctly
unique(df_ncro_final$Analyte)

rm(df_ncro, df_ndfs, df_ncro_new, df_sdtb, df_rs, df_edb)
```

Check DateTimes

```{r}
# any missing dates?
#df_ncro_final %>% filter(is.na(Date_Time)) %>% View()

# any missing times?
#df_ncro_final %>%
  #filter(format(Date_Time, '%H:%M:%S') == '00:00:00') %>%
  #distinct(`Date_Time`,`Sample Code`) %>% View()
```

### 5 - AEU (WDL)

Import AEU data (from WDL)

```{r}
fp_aeu_wdl <- abs_nutrient_path("AEU Data_WDL.csv")
df_aeu_wdl <- read_csv(fp_aeu_wdl)

rm(fp_aeu_wdl) #remove file paths
```

Filter for the years we want (AEU's EDI data only goes through 2021, so we need to get the 2022-2024 data from WDL)

```{r}
df_wdl_new <- df_aeu_wdl %>% 
  mutate(Date_Time = mdy_hm(`Collection Date`)) %>% #format datetime column
  mutate(year = year(Date_Time)) %>% #add a year column
  filter(year > 2021) #get rid of anything prior to 2022

range(df_wdl_new$Date_Time) #confirm dates are now from Jan 2022 - Jun 2024
```

Check/Format data

```{r}
unique(df_wdl_new$Matrix) #only shows Water, Natural
unique(df_wdl_new$`Sample Type`) #only shows Normal Samples 
unique(df_wdl_new$Analyte) #shows additional analytes that we need to filter out


df_aeu_wdl_final <- df_wdl_new %>%
  filter(Analyte == "Dissolved Ammonia" | Analyte == "Dissolved Nitrate + Nitrite" | Analyte == "Dissolved Organic Nitrogen" | Analyte == "Dissolved ortho-Phosphate" | Analyte == "Total Kjeldahl Nitrogen" | Analyte == "Total Phosphorus") %>% #only keep the analytes we want
  select(c("Long Station Name", "Short Station Name", "Station Number", "Date_Time", "Sample Code", "Analyte", "Result", "Rpt Limit", "Units", "Method", "Lab Dup", "Sample Type", "Description", "Notes")) %>% #remove unwanted columns
  add_column(Project = "Yolo Bypass Fish Monitoring Program") %>% #add Project column and have it all say YBFMP
  filter(!(is.na(`Result`) | `Result` == "N.S.")) #get rid of rows with NA or NS in the result column

rm(df_wdl_new) #remove intermediate data frames (need to keep raw WDL data, df_aeu_wdl, for later)
```

Check DateTimes

```{r}
# any missing dates?
#df_aeu_wdl_final %>% filter(is.na(Date_Time)) %>% View()

# any missing times?
#df_aeu_wdl_final %>%
  #filter(format(Date_Time, '%H:%M:%S') == '00:00:00') %>%
  #distinct(`Date_Time`,`Sample Code`) %>% View()
```

### 6 - WQA

Import WQA discrete data file

```{r}
fp_wqa <- abs_nutrient_path("WQA Discrete Data.csv")
df_wqa <- read_csv(fp_wqa, guess_max = 20000)
#problems(df_wqa)

rm(fp_wqa) #remove file paths
```

Check/format data

```{r}
unique(df_wqa$Matrix) #only shows Water, Natural
unique(df_wqa$`SampleType`) #only shows Normal Samples
unique(df_wqa$Analyte) #has dissolved TKN, which is typically used by the lab as a check (can remove)

df_wqa_final <- df_wqa %>%
  mutate(Date_Time = mdy_hm(`CollectionDate`)) %>% #format date
  filter(Analyte == "Dissolved Ammonia" | Analyte == "Dissolved Nitrate + Nitrite" | Analyte == "Dissolved Nitrate" | Analyte == "Dissolved Organic Nitrogen" | Analyte == "Dissolved ortho-Phosphate" | Analyte == "Total Kjeldahl Nitrogen" | Analyte == "Total Phosphorus") %>% #only keep the analytes we want
  select(c("LongStationName", "ShortStationName", "StationNumber", "Date_Time", "SampleCode", "Analyte", "Result", "RptLimit", "Units", "Method", "LabDup", "SampleType", "Description", "Notes")) %>% #remove unwanted columns
  rename("Short Station Name" = "ShortStationName", "Long Station Name" = "LongStationName", "Station Number" = "StationNumber", "Sample Code" = "SampleCode", "Rpt Limit" = "RptLimit", "Lab Dup" = "LabDup", "Sample Type" = "SampleType") %>% #rename columns to match other WDL data files
  add_column(Project = "WQA Discrete Sampling and Special Studies") %>% #add Project column
  filter(!(is.na(`Result`) | `Result` == "N.S.")) #get rid of rows with NA or NS in the result column

rm(df_wqa) #remove intermediate data frames
```

Check DateTimes

```{r}
# any missing dates?
#df_wqa_final %>% filter(is.na(Date_Time)) %>% View()

# any missing times?
#df_wqa_final %>%
  #filter(format(Date_Time, '%H:%M:%S') == '00:00:00') %>%
  #distinct(`Date_Time`,`Sample Code`) %>% View()
# a few missing times
```

## Non-WDL Data

### 1 - AEU (EDI)

Import AEU data from EDI

```{r}
#manual import
#fp_aeu_edi <- abs_nutrient_path("AEU Data_EDI.csv")
#df_aeu_edi <- read_csv(fp_aeu_edi)

df_aeu_edi <- get_edi_file(1391, "nutr_wq_clean") #1391 is the EDI package ID, nutr_wq_clean is the file name
```

Format columns and pivot (need to switch from wide to long format)

```{r}
df_aeu_new <- df_aeu_edi %>% 
  # mutate(Date_Time = ymd_hms(datetime)) %>% #format datetime column 
  # Dave's note: converting the datetime column isn't necessary since its already formatted as a
  # datetime. We just need to rename it so it matches the other Date_Time columns
  rename(Date_Time = datetime) %>% 
  mutate(year = year(Date_Time)) %>% #add a year column
  filter(year > 2009) #get rid of anything prior to 2010

#six date times failed to parse.
# df_aeu_test <- df_aeu_edi %>% 
#   mutate(Date_Time = ymd_hms(datetime)) %>%
#   filter(is.na(Date_Time))
#looks like just a few ones that didn't have time recorded, Date still looks good.
#So use date for Date_Time in those cases.
#Morgan: updated the mutate date line above to account for the 6 rows that didn't have times

# df_aeu_new <- df_aeu_edi %>% 
#   mutate(Date_Time = ymd_hms(datetime),  #format datetime column
#          Date_Time = case_when(is.na(Date_Time) ~ sample_date,
#                                TRUE ~ Date_Time),
#   year = year(Date_Time)) %>% #add a year column
#   filter(year > 2009) #get rid of anything prior to 2010

range(df_aeu_wdl_final$Date_Time)
range(df_aeu_new$Date_Time)
# There is some overlap between the WDL and EDI data - maybe these duplicates will be removed later?

#only keep the columns/anlaytes we want (and their respective sign columns) - remove the rest
df_aeu_new <- df_aeu_new %>%
  select(c("station_code", "event_id", "lab_comments", "diss_ammonia_sign", "diss_nitrate_nitrite_sign", "diss_orthophos_sign", "Date_Time", "year", "diss_ammonia", "diss_nitrate_nitrite", "don", "diss_orthophos", "tkn", "tot_phos")) %>%
  rename("Station_Name" = "station_code", "Sample Code" = "event_id") #rename station and sample ID columns

# df_aeu_long <- pivot_longer(df_aeu_new, cols = c("diss_ammonia", "diss_nitrate_nitrite", "diss_orthophos", "tot_phos", "don", "tkn"), names_to = "Analyte", values_to = "Result", values_drop_na = TRUE) #pivot the six analyte columns we want to keep

# Dave's tip: I would suggest replacing all of the pivoting and code separating each analyte with
  # the following more succinct and efficient code. Let me know if you want me to go over how this
  # works.
# Pivot the six analyte columns we want to keep to longer data structure
df_aeu_comb <- df_aeu_new %>% 
  # Add a value suffix to the columns that contain nutrient values
  rename_with(
    \(x) paste0(x, "_value"), 
    c(diss_ammonia, diss_nitrate_nitrite, don, diss_orthophos, tkn, tot_phos)
  ) %>% 
  # Pivot analytes longer, creating a sign and value column for each analyte
  pivot_longer(
    cols = ends_with(c("sign", "value")),
    names_to = c("Analyte", ".value"),
    names_pattern = "(.*)_(.*)",
    values_drop_na = TRUE
  ) %>% 
  mutate(
    # Create a Detection Condition column and if there's a "<" in the sign column, make it say Not
      # Detected, otherwise say Detected
    Detection_Condition = case_match(
      sign, 
      "<" ~ "Not Detected", 
      c("=", NA_character_) ~ "Detected"
    ),
    # Create an RL column and make it say the result if the detection column says "Not Detected",
      # otherwise leave as NA
    "Rpt Limit" = if_else(Detection_Condition == "Not Detected", value, NA_real_),
    # If the value is not detected, change result to NA, otherwise leave the result
    Result = if_else(Detection_Condition == "Not Detected", NA_real_, value),
    # Remove the unwanted columns used within this mutate
    .keep = "unused"
  ) %>% 
  # Rename all analyte names to match WDL format. Recode is superseeded by case_match, using that
    # function instead
  mutate(
    Analyte = case_match(
      Analyte, 
      "diss_ammonia" ~ 'Dissolved Ammonia',
      "diss_nitrate_nitrite" ~ 'Dissolved Nitrate + Nitrite',
      "diss_orthophos" ~ 'Dissolved ortho-Phosphate',
      "tkn" ~ 'Total Kjeldahl Nitrogen', 
      "don" ~ 'Dissolved Organic Nitrogen', 
      "tot_phos" ~ 'Total Phosphorus'
    )
  )
```

For analytes with non-detects: split into individual data frames, format, then recombine - not necessary if you use the code above.

```{r}
# #Ammonia
# df_aeu_ammonia <- df_aeu_long %>%
#   filter(Analyte=='diss_ammonia') %>% #filter for the analyte we want
#   select(-c(diss_nitrate_nitrite_sign, diss_orthophos_sign)) #remove the unwanted analyte sign columns
# 
# df_aeu_ammonia <- df_aeu_ammonia %>%
#   mutate(Detection_Condition = case_when(str_detect(diss_ammonia_sign, '<') ~ "Not Detected", #if there's a "<", then say Not Detected
#          .default = "Detected")) %>% #otherwise say Detected
#   mutate("Rpt Limit" = case_when(str_detect(Detection_Condition, 'Not Detected') ~ Result, #create a RL column and make it say the result if the detection column says "Not Detected"
#          .default = NA)) %>% #otherwise leave as NA
#   mutate(result2 = case_when(str_detect(Detection_Condition, 'Not Detected') ~ NA, #if the value is not detected, change result to NA
#          .default = Result)) %>% #otherwise leave the result
#   select(-c(Result, diss_ammonia_sign)) %>% #remove the old result column and old sign column
#   rename(Result = result2) #rename to just say Result
# 
# #Nitrate Nitrite
# df_aeu_nitrate <- df_aeu_long %>%
#   filter(Analyte=='diss_nitrate_nitrite') %>% #filter for the analyte we want
#   select(-c(diss_ammonia_sign, diss_orthophos_sign)) #remove the unwanted analyte sign columns
# 
# df_aeu_nitrate <- df_aeu_nitrate %>%
#   mutate(Detection_Condition = case_when(str_detect(diss_nitrate_nitrite_sign, '<') ~ "Not Detected", #if there's a "<", then Not Detected
#          .default = "Detected")) %>% #otherwise say Detected
#   mutate("Rpt Limit" = case_when(str_detect(Detection_Condition, 'Not Detected') ~ Result, #create a RL column and make it say the result if the detection column says "No"
#          .default = NA)) %>%
#   mutate(result2 = case_when(str_detect(Detection_Condition, 'Not Detected') ~ NA, #if the value is not detected, change result to NA
#          .default = Result)) %>% #otherwise leave the result
#   select(-c(Result, diss_nitrate_nitrite_sign)) %>% #remove the old result column and old sign column
#   rename(Result = result2) #rename to just say Result
# 
# #Orthophosphate
# df_aeu_ortho <- df_aeu_long %>%
#   filter(Analyte=='diss_orthophos') %>% #filter for the analyte we want
#   select(-c(diss_ammonia_sign, diss_nitrate_nitrite_sign)) #remove the unwanted analyte detection columns
# 
# df_aeu_ortho <- df_aeu_ortho %>%
#   mutate(Detection_Condition = case_when(str_detect(diss_orthophos_sign, '<') ~ "Not Detected", #if there's a "<", then Not Detected
#          .default = "Detected")) %>% #otherwise say Detected
#   mutate("Rpt Limit" = case_when(str_detect(Detection_Condition, 'Not Detected') ~ Result, #create a RL column and make it say the result if the detection column says "No"
#          .default = NA)) %>%
#   mutate(result2 = case_when(str_detect(Detection_Condition, 'Not Detected') ~ NA, #if the value is not detected, change it to NA
#          .default = Result)) %>% #otherwise leave the result
#   select(-c(Result, diss_orthophos_sign)) %>% #remove the old result column and old sign column
#   rename(Result = result2) #rename to just say Result
# 
# #combine three analyte data frames back together
# df_aeu_NDs <- bind_rows(df_aeu_ammonia, df_aeu_nitrate, df_aeu_ortho)
# 
# #rename all analyte names to match WDL format
# df_aeu_NDs <- df_aeu_NDs %>%
#   # recode is superseeded by case_match, using that function instead
#   mutate(
#     analyte2 = case_match(
#       Analyte, 
#       "diss_ammonia" ~ 'Dissolved Ammonia',
#       "diss_nitrate_nitrite" ~ 'Dissolved Nitrate + Nitrite',
#       "diss_orthophos" ~ 'Dissolved ortho-Phosphate' )
#   ) %>%
#   select(-Analyte) %>% #remove old analyte column
#   rename(Analyte = analyte2) #rename new analyte column
# 
# rm(df_aeu_ammonia, df_aeu_nitrate, df_aeu_ortho) #remove intermediate data frames
```

```{r}
# # check if any data was lost
# df_aeu_test <- df_aeu_long %>%
#   filter(Analyte %in% c('diss_ammonia','diss_nitrate_nitrite','diss_orthophos'))
# 
# nrow(df_aeu_NDs) == nrow(df_aeu_test)
# 
# rm(df_aeu_test)
```

Add data for the three non-detect analytes back into df_aeu_long - not necessary if you use the code above.

```{r}
# df_aeu_long2 <- df_aeu_long %>%
#   filter(!(Analyte == "diss_ammonia" | Analyte == "diss_nitrate_nitrite" | Analyte == "diss_orthophos")) %>% #remove data for the three analytes with NDs
#   select(-c(diss_ammonia_sign, diss_nitrate_nitrite_sign, diss_orthophos_sign)) %>% #remove old sign columns
#   add_column(Detection_Condition = "Detected") #add the detection condition column and have it all say detected, since there are no NDs
# 
# df_aeu_comb <- bind_rows(df_aeu_NDs, df_aeu_long2) #combine data frames together
# 
# df_aeu_comb <- df_aeu_comb %>%
#   mutate(
#     analyte2 = case_match(
#       Analyte, 
#       "tkn" ~ 'Total Kjeldahl Nitrogen', 
#       "don" ~ 'Dissolved Organic Nitrogen', 
#       "tot_phos" ~ 'Total Phosphorus',
#       .default = Analyte
#     )
#   ) %>% #rename remaining analyte names to match WDL names
#   select(-Analyte) %>% #remove old analyte column
#   rename(Analyte = analyte2) #rename new analyte column
#   
# unique(df_aeu_comb$Analyte) #confirm the analyte names are correct
# 
# rm(df_aeu_NDs, df_aeu_long, df_aeu_long2, df_aeu_new) #remove intermediate data frames
```

Determine analytical methods used (from WDL data) and add into EDI data set

```{r}
# Dave's tip: I would suggest replacing all of the code used to define analytical methods with the
# following more succinct and efficient code. Let me know if you want me to go over how this works.
df_methods <- df_aeu_wdl %>% 
  distinct(Analyte, Method) %>% 
  arrange(Analyte)

# Methods used:
# Dissolved Ammonia - only EPA 350.1 (DWR modified and not modified)
# Dissolved Nitrate + Nitrite - only Std Method 4500-NO3-F (DWR modified and not modified)
# Total Kjeldahl Nitrogen - only EPA 351.2 (DWR modified and not modified)
# Total Phosphorus - only EPA 365.4 (DWR modified and not modified)
# Dissolved Organic Nitrogen - only EPA 351.2/EPA 350.1 (D) (DWR modified and not modified)
# Dissolved ortho-Phosphate - only EPA 365.1 (DWR modified and not modified)

# df_ammonia <- df_aeu_wdl %>%
#   filter(Analyte=='Dissolved Ammonia') #create a subset of data for just dissolved ammonia
# 
# unique(df_ammonia$Method) #look at the method/s used, only EPA 350.1 (DWR modified and not modified)
# 
# df_nitratenitrite <- df_aeu_wdl %>%
#   filter(Analyte=='Dissolved Nitrate + Nitrite') #create a subset of data for just dissolved ammonia
# 
# unique(df_nitratenitrite$Method) #look at the method/s used, only Std Method 4500-NO3-F (DWR modified and not modified)
# 
# df_tkn <- df_aeu_wdl %>%
#   filter(Analyte=='Total Kjeldahl Nitrogen') #create a subset of data for just TKN
# 
# unique(df_tkn$Method) #look at the method/s used, only EPA 351.2 (DWR modified and not modified)
# 
# df_phos <- df_aeu_wdl %>%
#   filter(Analyte=='Total Phosphorus') #create a subset of data for just total phosphorus
# 
# unique(df_phos$Method) #look at the method/s used, only EPA 365.4 (DWR modified and not modified)
# 
# df_don <- df_aeu_wdl %>%
#   filter(Analyte=='Dissolved Organic Nitrogen') #create a subset of data for just DON
# 
# unique(df_don$Method) #look at the method/s used, only EPA 351.2/EPA 350.1 (D) (DWR modified and not modified)
# 
# df_ortho <- df_aeu_wdl %>%
#   filter(Analyte=='Dissolved ortho-Phosphate') #create a subset of data for just dissolved orthophos
# 
# unique(df_ortho$Method) #look at the method/s used, only EPA 365.1 (DWR modified and not modified)
# 
# rm(df_ammonia, df_nitratenitrite, df_tkn, df_phos, df_don, df_ortho) #remove intermediate data frames
```

Add methods to EDI data set

```{r}
# Update data frame with analyte and method info by removing duplicates, revising method names slightly
# df_methods2 <- data.frame(Analyte = c("Dissolved Ammonia", "Dissolved Nitrate + Nitrite", "Total Kjeldahl Nitrogen", "Total Phosphorus", "Dissolved Organic Nitrogen", "Dissolved ortho-Phosphate"),
#               Method = c("EPA 350.1","Std Method 4500-NO3-, F", "EPA 351.2", "EPA 365.4", "EPA 351.2/EPA 350.1", "EPA 365.1"))

df_methods <- df_methods %>% 
  filter(!str_detect(Method, "DWR Modified")) %>% 
  mutate(
    Method = str_trim(str_remove_all(Method, "\\(D\\)|\\[1\\]\\*")),
    Method = if_else(Analyte == "Dissolved Nitrate + Nitrite", "Std Method 4500-NO3-, F", Method)
  )

df_aeu_comb <- left_join(df_aeu_comb, df_methods, by = "Analyte") #use left join to add methods to EDI data based on Analyte

# double check no NAs
#df_aeu_comb %>% filter(is.na('Method')) %>% View()
```

Add WDL station numbers to data (need this to get lat/longs added later)

```{r}
unique(df_aeu_comb$Station_Name) #confirm which stations are in the data file 

df_aeu_station_numbers <- data.frame(Station_Name = c("SHR", "STTD", "LIS"),
                                    Station_Number = c("A0200000","A0D82120386", "B9D82851352"))

df_aeu_comb <- left_join(df_aeu_comb, df_aeu_station_numbers, by = "Station_Name")

df_aeu_comb <- df_aeu_comb %>%
  rename("Station Number" = Station_Number) #rename the station number column to match the WDL data

# double check no NAs
#df_aeu_comb %>% filter(is.na('Station Number')) %>% View()

rm(df_aeu_station_numbers)
```

Add sample codes from WDL data

```{r}
df_sample_codes <- df_aeu_wdl %>%
  select(c("Station Number", "Sample Code", "Collection Date", "Analyte")) %>% #only keep the columns we need
  mutate(Date_Time = mdy_hm(`Collection Date`), .keep = "unused") %>% #format date and remove old date time column
  # select(-`Collection Date`)
  distinct() #get rid of repeated rows

df_aeu_comb <- left_join(df_aeu_comb, df_sample_codes, by = c("Station Number", "Date_Time", "Analyte"))

df_aeu_comb <- df_aeu_comb %>%
  select(-`Sample Code.x`) %>% #remove old sample code column
  rename("Sample Code" = `Sample Code.y`) #rename new sample code column

rm(df_sample_codes)
```

Add reporting limits to "Detected" data from WDL

```{r}
df_RLs <- df_aeu_wdl %>%
  select(c("Station Number", "Sample Code", "Collection Date", "Analyte", "Rpt Limit")) %>%
  mutate(Date_Time = mdy_hm(`Collection Date`), .keep = "unused") %>% #format date
  # select(-`Collection Date`)
  distinct()

df_detected <- df_aeu_comb %>%
  filter(Detection_Condition == "Detected") %>% #filter for just the detected values
  select(-`Rpt Limit`) #remove RL column

# unique(df_detected$`Rpt Limit`) #confirmed there are no RLs before removing the RL column above

df_detected <- left_join(df_detected, df_RLs, by = c("Station Number", "Date_Time", "Sample Code", "Analyte"))

df_not_detected <- df_aeu_comb %>%
  filter(Detection_Condition == "Not Detected") #filter for just the not detected values

df_aeu_comb <- bind_rows(df_detected, df_not_detected)

# check everything was added
nrow(df_aeu_comb) == nrow(df_detected)+nrow(df_not_detected)

rm(df_detected, df_not_detected, df_RLs)
```

Format columns to match WDL data 

```{r}
df_aeu_edi_final <- df_aeu_comb %>%
  select(-year) %>% #remove the year column
  rename("Notes" = lab_comments, "Short Station Name" = Station_Name) %>% #rename the lab comments and station column to match WDL
  add_column(Units = "mg/L") %>% #add a Units column and have all say mg/L (which is what metadata says on EDI, but will update later based on methods)
  add_column(Project = "Yolo Bypass Fish Monitoring Program") #add Project column and have it all say YBFMP

range(df_aeu_edi_final$Date_Time) #confirm data is from 2013-2021

rm(df_aeu_comb, df_aeu_wdl, df_aeu_edi, df_aeu_new) #remove intermediate data frames
```

### 2 - DEMP

Import DEMP data and LSZ station lat/longs

```{r}
fp_demp <- abs_nutrient_path("DEMP Data.csv")
df_demp <- read_csv(fp_demp)

fp_latlong <- abs_nutrient_path("DEMP_LSZ_LatLong.csv")
df_latlong <- read_csv(fp_latlong)

rm(fp_demp, fp_latlong) #remove file paths
```

Check data

```{r}
unique(df_demp$`Observed Property ID`) #confirm it includes only the six analytes we want
unique(df_demp$`Lab: Analysis Method`) #look at list of analytical methods
unique(df_demp$`QC: Type`) #shows only NA, which means no blanks or dups are present
range(df_demp$`Observed DateTime`) #confirm dates are from 2010-2024
```

Get LSZ station lat/longs added to lat/long column

```{r}
df_LSZ <- df_demp %>%
  filter(`Location ID` == "LSZ2" | `Location ID` == "LSZ6" | `Location ID` == "LSZ2-SJR" | `Location ID` == "LSZ6-SJR") %>% #keep LSZs only
  select(-c(Latitude, Longitude)) #remove lat and long columns

df_latlong <- df_latlong %>%
  select(c("Location", "Activity ID", "Latitude", "Longitude")) %>% #remove unwanted columns
  rename("Location ID" = "Location", "Activity Name" = "Activity ID") #rename columns to match df_DEMP

df_LSZ <- left_join(df_LSZ, df_latlong, by = c("Activity Name", "Location ID")) #add the lat/longs to the LSZ data

df_demp <- df_demp %>%
  filter(!(`Location ID` == "LSZ2" | `Location ID` == "LSZ6" | `Location ID` == "LSZ2-SJR" | `Location ID` == "LSZ6-SJR")) #remove the LSZs from the original DEMP data frame

unique(df_demp$`Location ID`) #confirm the LSZs are no longer in the data set

df_demp <- bind_rows(df_demp, df_LSZ) #add LSZ data back into original data file
```

Format columns

```{r}
df_demp_new <- df_demp %>%
  mutate(Date_Time = ymd_hms(`Observed DateTime`)) %>% #format date
  mutate(Detection_Condition = case_when(str_detect(`Lab: Detection Condition`, "NOT_DETECTED") ~ "Not Detected", #fix capitalization
         .default = "Detected")) %>% #otherwise say Detected
  select(c("Location ID", "Observed Property ID", "Date_Time", "Result Value", "Result Unit", "Activity Name", "Field: Comment", "Lab: Analysis Method", "Detection_Condition", "Lab: MDL", "Lab: MRL", "Lab: Comment", "Latitude", "Longitude", "EA_Reports To")) %>% #remove unwanted rows
  add_column(Project = "EMP Discrete Water Quality Monitoring") %>% #add Project column and have it all say EMP DWQ Monitoring
  rename(Station_Name = "Location ID", Analyte = "Observed Property ID", Result = "Result Value", Units = "Result Unit", "Sample Code" = "Activity Name", "Rpt Limit" = "Lab: MRL") #rename columns to match WDL data
  

rm(df_demp, df_LSZ, df_latlong) #remove intermediate data frames
```

Clean up method and analyte names

```{r}
df_demp_methods <- df_demp_new %>%
  mutate(Method = str_remove(`Lab: Analysis Method`, "\\;.*"), .keep = "unused") #remove everything after and including the first ";"

#rename dissolved orthophosphate to match WDL data
df_demp_methods <- df_demp_methods %>%
  mutate(
    Analyte = if_else(Analyte == "Dissolved Orthophosphate", "Dissolved ortho-Phosphate", Analyte)
  )

# confirm the method names are the same as those used in other data sets
#df_methods %>% 
  #full_join(df_demp_methods %>% distinct(Analyte, Method), keep = TRUE) %>% 
  #drop_na(Analyte.y) %>% 
  #arrange(Analyte.y) %>% View()
# There are a few additional methods in the DEMP dataset not found in the others - not sure if this matters
```

Deal with the Test America data that reports to the MDL (change to RL)

```{r}
df_TA <- df_demp_methods %>%
  filter(Date_Time > "2019-10-01" & Date_Time < "2020-04-01") #filter dates between Oct 2019 and Mar 2020, when Test America was used

range(df_TA$Date_Time) #confirm correct time frame

df_TA <- df_TA %>%
  mutate(result_new = case_when((Result < `Rpt Limit`) ~ NA, #if the result is less than the RL, make it NA
         .default = Result)) %>% #otherwise leave the result as is
  mutate(detection_new = case_when((Result < `Rpt Limit`) ~ "Not Detected", #if the result is less than the RL, then add "Not Detected"
         .default = Detection_Condition)) %>% #otherwise leave as is
  select(-c(Result, Detection_Condition)) %>% #remove the old result/detection columns
  rename(Result = result_new, Detection_Condition = detection_new) #rename new columns

df_Bryte <- df_demp_methods %>%
  filter(!(Date_Time > "2019-10-01" & Date_Time < "2020-04-01")) #remove TA data

df_demp_final <- bind_rows(df_Bryte, df_TA)

df_demp_final <- df_demp_final %>%
  select(-c(`Lab: MDL`, `EA_Reports To`)) %>% #remove columns that are no longer needed
  rename(Description = "Lab: Comment", Notes = "Field: Comment") #rename columns to match WDL data

rm(df_demp_methods, df_demp_new, df_TA, df_Bryte) #remove intermediate data frames
```

### 3 - FRP

Import FRP data

```{r}
#manual import
#fp_frp <- abs_nutrient_path("FRP Data.csv")
#df_frp <- read_csv(fp_frp)

df_frp <- get_edi_file(269, "nutrients_FRP2023") #269 is the EDI package ID, nutrients_FRP2023 is the file name
```

Check data

```{r}
unique(df_frp$`Analyte`) #there are some additional analytes we will need to filter out
unique(df_frp$`SampleType`) #normal samples, blank samples, replicate/duplicate samples

#analysis method column is not present, will need to add those in
#some collection times are missing
#some rows are missing station names - remove these
#missing lat/longs - need to get these from CDFW
#will need to combine date and time columns
```

Add analytical methods

```{r}
df_frp_nutrients <- df_frp %>%
  filter(Analyte == "Dissolved Ammonia" | Analyte == "Dissolved Nitrate + Nitrite" | Analyte == "Dissolved Organic Nitrogen" | Analyte == "Dissolved ortho-Phosphate" | Analyte == "Total Kjeldahl Nitrogen" | Analyte == "Total Phosphorus")  %>% #only keep the analytes we want
  filter(SampleType == 'Normal Sample') # filter out QA samples

#check rows with missing stations - 45 rows that will need to be removed
df_NA_test <- df_frp_nutrients %>% filter(is.na(Location))

df_frp_nutrients <- df_frp_nutrients %>%
  filter(!is.na(Location))

df_frp_methods <- left_join(df_frp_nutrients, df_methods, by = 'Analyte')

#confirmed these were the methods used by FRP, based on the SOP on EDI
```

Format/filter data

```{r}
df_frp_new <- df_frp_methods %>%
  # mutate(Date = ymd(Date)) %>% #format the date column - not necessary, column is already formatted as a date
  unite(col = "Date_Time", Date, StartTime, sep = " ", na.rm=TRUE) #combine date time columns, remove any times that were NA

#Date_Time column is a character, need to format but there's a problem with having some of the times missing (can't format everything all at once) - use the parse_date_time function for this
df_frp_new <- df_frp_new %>%
  mutate(Date_Time = parse_date_time(`Date_Time`, orders = c("Ymd HMS", "Ymd"))) #format date, include both formats since some times missing

range(df_frp_new$`Date_Time`) #date range is 2018-2023

str(df_frp_new) #confirm the date time column is formatted as a date
  
df_frp_final <- df_frp_new %>%
  rename("Short Station Name" = "Location", "Sample Code" = "SampleID_frp", "Rpt Limit" = "RptLimit", "Notes" = "Comments", "Sample Type" = "SampleType") %>%
  add_column("Long Station Name" = NA) %>% #add a long station name column to match WDL data 
  select(c("Long Station Name", "Short Station Name", "Date_Time", "Sample Code", "Analyte", "Result", "Rpt Limit", "Units", "Method", "Sample Type", "Notes")) %>% #remove unwanted columns
  add_column(Project = "Fish Restoration Program") %>% #add Project column and have it all say FRP
  filter(!(`Result` == "N.S.")) #get rid of "Not Sampled" rows

#missing Latitude, Longitude, so would need to add those to the list of columns kept above if adding in later
#missing lab dups and description column, so should probably combine the FRP data after lab dups have been filtered out of WDL data

# check no NA Date_Time
#df_frp_final %>% filter(is.na(Date_Time)) %>% View()

# any missing times?
#df_frp_final %>%
  #filter(format(Date_Time, '%H:%M:%S') == '00:00:00') %>%
  #distinct(`Date_Time`,`Sample Code`) %>% View()

# TODO: quite a few, but fine?

rm(df_frp, df_frp_new, df_frp_methods, df_frp_nutrients, df_NA_test)
```

## Data Integration and Formatting

### Combine WDL Data

Combine data sets from WDL (O&M, Tidal Parr, DISE-NDFS, NCRO, WQA, AEU-WDL)

```{r}
df_nutrients_WDL <- bind_rows(df_om_final, df_tp_final, df_dise_ndfs_final, df_ncro_final, df_aeu_wdl_final, df_wqa_final) #combine WDL data sets

unique(df_nutrients_WDL$Project) #confirm all data sets were combined

# check no data was lost
nrow(df_nutrients_WDL) == nrow(df_om_final)+nrow(df_tp_final)+nrow(df_dise_ndfs_final)+nrow(df_ncro_final)+nrow(df_aeu_wdl_final)+nrow(df_wqa_final)

rm(df_om_final, df_tp_final, df_dise_ndfs_final, df_ncro_final, df_aeu_wdl_final, df_wqa_final) #remove individual WDL data frames
```

### Lab Duplicates

Remove duplicates run by the laboratory

```{r}
df_nutrients_WDL <- df_nutrients_WDL %>%
  filter(is.na(`Lab Dup`)) #get rid of all the lab dups, which have a "Y" in the Lab Dup column

unique(df_nutrients_WDL$`Lab Dup`) #confirm that there are no more "Y"'s in the Lab Dup column

df_nutrients_WDL <- df_nutrients_WDL %>%
  select(-`Lab Dup`) #remove the Lab Dup column
```

### Method Names

Clean up method names

```{r}
# Dave's tip: I would suggest replacing all of the code used to check analytical methods with the
# following more succinct and efficient code. Let me know if you want me to go over how this works.
df_nutrients_WDL %>% 
  distinct(Analyte, Method) %>% 
  arrange(Analyte)

# Methods used:
# Dissolved Ammonia - only EPA 350.1 (DWR modified and not modified)
# Dissolved Nitrate + Nitrite - only EPA 353.2 (O&M data from TA) and Std Method 4500-NO3-, F (DWR
  # modified and not modified)
# Total Kjeldahl Nitrogen - only EPA 351.2 (DWR modified and not modified)
# Total Phosphorus - only EPA 365.4 (DWR modified and not modified)
# Dissolved Organic Nitrogen - only EPA 351.2/EPA 350.1 (D) (DWR modified and not modified)
# Dissolved ortho-Phosphate - only EPA 365.1 (DWR modified and not modified)
# Dissolved Nitrate (O&M only) - only EPA 300.0 (some of it says 28 day hold, but applies to all data)

# df_ammonia <- df_nutrients_WDL %>%
#   filter(Analyte=='Dissolved Ammonia') #create a subset of data for just dissolved ammonia
# 
# unique(df_ammonia$Method) #look at the method/s used, only EPA 350.1 (DWR modified and not modified)
# 
# df_nitratenitrite <- df_nutrients_WDL %>%
#   filter(Analyte=='Dissolved Nitrate + Nitrite') #create a subset of data for just dissolved nitrate + nitrite
# 
# unique(df_nitratenitrite$Method) #look at the method/s used, only EPA 353.2 (O&M data from TA) and Std Method 4500-NO3-, F (DWR modified and not modified)
# 
# df_tkn <- df_nutrients_WDL %>%
#   filter(Analyte=='Total Kjeldahl Nitrogen') #create a subset of data for just TKN
# 
# unique(df_tkn$Method) #look at the method/s used, only EPA 351.2 (DWR modified and not modified)
# 
# df_phos <- df_nutrients_WDL %>%
#   filter(Analyte=='Total Phosphorus') #create a subset of data for just total phosphorus
# 
# unique(df_phos$Method) #look at the method/s used, only EPA 365.4 (DWR modified and not modified)
# 
# df_don <- df_nutrients_WDL %>%
#   filter(Analyte=='Dissolved Organic Nitrogen') #create a subset of data for just DON
# 
# unique(df_don$Method) #look at the method/s used, only EPA 351.2/EPA 350.1 (D) (DWR modified and not modified)
# 
# df_ortho <- df_nutrients_WDL %>%
#   filter(Analyte=='Dissolved ortho-Phosphate') #create a subset of data for just dissolved orthophos
# 
# unique(df_ortho$Method) #look at the method/s used, only EPA 365.1 (DWR modified and not modified) and Std Method 4500-P, F
# 
# df_nitrate <- df_nutrients_WDL %>%
#   filter(Analyte=='Dissolved Nitrate') #create a subset of data for just dissolved nitrate (O&M only)
# 
# unique(df_nitrate$Method) #look at the method/s used, only EPA 300.0 (some of it says 28 day hold, but applies to all data)

unique(df_nutrients_WDL$Method)
unique(df_nutrients_WDL$Analyte)

df_nutrients_WDL <- df_nutrients_WDL %>%
  mutate(
    methods2 = case_match(
      Method,
      c("EPA 350.1 (DWR Modified) [1]*", "EPA 350.1 (D) [1]*") ~ 'EPA 350.1',
      c("Std Method 4500-NO3-F (DWR Modified) [1]*", "Std Method 4500-NO3-F [1]*") ~ 'Std Method 4500-NO3-, F',
      c("EPA 351.2/EPA 350.1 (D) (DWR Modified) [1]*", "EPA 351.2/EPA 350.1 (D) [1]*") ~ 'EPA 351.2/EPA 350.1',
      c("EPA 365.4 (DWR Modified) [1]*", "EPA 365.4 [1]*") ~ "EPA 365.4",
      c("EPA 351.2 (DWR Modified) [1]*", "EPA 351.2 [1]*") ~ "EPA 351.2",
      c("EPA 365.1 DWR Modified [1]*", "EPA 365.1 [1]*") ~ "EPA 365.1",
      c("EPA 300.0 28d Hold [1]*", "EPA 300.0 [1]*") ~ "EPA 300.0",
      "EPA 353.2 [1]*" ~ "EPA 353.2"
    )
  ) %>% #clean up method names (will call out modified methods in metadata)
  select(-Method) %>% #remove old method column
  rename(Method = methods2) #rename new method column

unique(df_nutrients_WDL$Method) #confirm they look the way we want

# rm(df_ammonia, df_nitratenitrite, df_tkn, df_phos, df_don, df_ortho, df_nitrate)
```

### Add FRP Data (Non-WDL)

Adding now since it didn't have lab dups, and methods were clean, but still has field dups/blanks that need to be removed)

```{r}
names(df_frp_final) %in% names(df_nutrients_WDL)

df_nutrients_WDL <- bind_rows(df_nutrients_WDL, df_frp_final)

rm(df_frp_final)
```

### Field Dups/Blanks

Remove everything besides normal samples

```{r}
unique(df_nutrients_WDL$`Sample Type`)

df_nutrients_WDL <- df_nutrients_WDL %>%
  filter(`Sample Type` == "Normal Sample") %>% #get rid of everything besides normal samples
  select(-`Sample Type`) #remove the sample type column

# Perry: I also removed most of these as it went through, but still good to check here. I like doing it for each individual dataframe as well just in case there's some weird edge cases that are hard to catch once it's synthesized.
```

### Non-Detects

Deal with non-detect data

```{r}
df_nutrients_WDL <- df_nutrients_WDL %>%
  mutate(Detection_Condition = case_when(str_detect(Result, '<') ~ 'Not Detected', #create a detection column and make it say "Not Detected" if the result column contains a "<" symbol
         .default = "Detected")) %>% #otherwise say Detected
  mutate(result2 = case_when(str_detect(Result, '<') ~ NA, #create a new result column and make it remove the value if the result column contains a "<" symbol
         .default = Result)) %>% #otherwise leave the result as is
  select(-Result) %>% #remove the old result column
  rename(Result = result2) #rename new result column

#deal with O&M issue where results are 0, but should be ND (contain J flag, which means MDL < result < RL)

#first look at the sample with results = 0
# df_nutrients_WDL %>%
#   filter(Result == 0) %>%
#   View()

df_nutrients_WDL <- df_nutrients_WDL %>%
  mutate(Detection_Condition = case_when((Result == 0) ~ "Not Detected", #when a result is 0, change the detection condition to "not detected"
         .default = Detection_Condition)) %>% #otherwise leave the detection column as is
  mutate(Result = case_when((Result == 0) ~ NA, #when a result is 0, change to NA
         .default = Result)) #otherwise leave result as is

#confirm the J flags were converted to NA, Not Detected
df_nutrients_WDL %>%
  filter(Notes == "J") %>%
  View()
```

### Add AEU Data (Non-WDL)

Check which values would be converted to NA

```{r}
check_non_numeric(df_nutrients_WDL, 'Result')

check_non_numeric(df_aeu_edi_final, 'Result')
```

Adding now since there were no lab/field dups, or blanks, the method names were clean, and the non-detects were already fixed 

```{r}
names(df_aeu_edi_final) %in% names(df_nutrients_WDL)

#need to change Result columns to numeric before combining

# mutate_at is superseeded - its easy enough to use a simple mutate here
df_nutrients_WDL <- df_nutrients_WDL %>%
  # mutate_at('Result', as.numeric) #convert to numeric
  mutate(Result = as.numeric(Result))

# The Result column in df_aeu_edi_final is already numeric
# df_aeu_edi_final <- df_aeu_edi_final %>%
#   mutate_at('Result', as.numeric) #convert to numeric

df_nutrients_comb <- bind_rows(df_nutrients_WDL, df_aeu_edi_final)

rm(df_aeu_edi_final, df_nutrients_WDL)
```

### Add Lat/Longs & Update Station Names 

Use the master station list (duplicate stations removed) to add the lat/longs

```{r}
fp_stations_discrete <- abs_nutrient_path("Discrete_Stations_Consolidated.csv")
df_stations_discrete <- read_csv(fp_stations_discrete)

df_stations_discrete <- df_stations_discrete %>%
  select(c(Station_Name, Latitude, Longitude, `Station Number`)) #remove unwanted columns

#Checks
df_name_check <- df_stations_discrete %>%
  count(Station_Name) %>%
  filter(n > 1) #Lindsey Slough is the only duplicated station name, one is from FRP, which doesn't have a lat/long associated with it

df_gps_check <- df_stations_discrete %>%
  count(Latitude, Longitude) %>%
  filter(n > 1) #found two stations with lat/long of 38.24285, -121.6836 - fixed!

df_station_dup <- df_stations_discrete %>%
  filter(Latitude == 38.24285 & Longitude == -121.6836) 
#stations with the same GPS coords are South Tip of Liberty Island (WQA) and Liberty at S End (DISE NDFS), which have different station numbers

df_number_check <- df_stations_discrete %>%
  count(`Station Number`) %>%
  filter(n > 1) #no duplicated station numbers (just NAs)

#Add lat/longs
df_frp <- df_nutrients_comb %>%
  filter(is.na(`Station Number`)) %>% #need to first separate FRP data since there are no station numbers to match
  rename(Station_Name = "Short Station Name")

unique(df_frp$Project) #confirm it's all FRP data that got taken out

df_no_frp <- df_nutrients_comb %>%
  filter(!(is.na(`Station Number`))) #remove the FRP data from the main data frame

df_no_frp_latlong <- left_join(df_no_frp, df_stations_discrete, by = "Station Number") #add station info

df_nutrients_comb <- bind_rows(df_no_frp_latlong, df_frp) #add FRP data back in

df_nutrients_comb <- df_nutrients_comb %>%
  select(-c(`Long Station Name`, `Short Station Name`)) #get rid of WDL station name columns

rm(df_frp, df_no_frp, df_no_frp_latlong, df_station_dup, df_gps_check, df_name_check, df_number_check) #remove intermediate data frames
rm(fp_stations_discrete) #remove file paths
```

### Add DEMP Data (Non-WDL)

Adding now since no lab/field dups or blanks, method names were clean, non-detects were fixed, and lat/longs were present

```{r}
#station numbers were missing from the DEMP data, so we need to add those first

df_station_numbers <- df_stations_discrete %>%
  select(-c(Latitude, Longitude)) #remove lat/long columns since we don't need them for the DEMP data

df_demp_final <- left_join(df_demp_final, df_station_numbers, by = "Station_Name") #add the station numbers based on station name

names(df_demp_final) %in% names(df_nutrients_comb) #confirm they have the same columns

# The Result column in df_aeu_edi_final is already numeric
# df_demp_final <- df_demp_final %>%
  # mutate_at('Result', as.numeric)  #convert DEMP result column to numeric

#When you converted Longitude to numeric NA's happened, so this is to figure out why
df_nutrients_combx <- df_nutrients_comb %>%
  mutate(Lat1 = as.numeric(Latitude)) %>% #convert latitude column to numeric
  mutate(Long1 = as.numeric(Longitude)) %>% #convert longitude column to numeric
  filter(is.na(Long1) & !is.na(Longitude)) 

unique(df_nutrients_combx$Longitude)
#OH! THere is a space at the end of 'longitude' at station BET

df_nutrients_comb <- df_nutrients_comb %>%
  mutate(Latitude = as.numeric(str_trim(Latitude))) %>% #convert latitude column to numeric, but trim white space first
  mutate(Longitude = as.numeric(str_trim(Longitude))) #convert longitude column to numeric, but trim white space first
#no NAs!

df_nutrients_comb <- bind_rows(df_nutrients_comb, df_demp_final)

unique(df_nutrients_comb$Project) #confirm all projects are included

rm(df_demp_final, df_station_numbers, df_stations_discrete, df_nutrients_combx)
```

### Units

Check units

```{r}
# Dave's tip: I would suggest replacing all of the code used to check units and analytical methods
# with the following more succinct and efficient code. Let me know if you want me to go over how
# this works.
df_nutrients_comb %>% 
  distinct(Analyte, Method, Units) %>% 
  arrange(Analyte, Method, Units) %>% View()

# #dissolved ammonia
# df_ammonia <- df_nutrients_comb %>%
#   filter(Analyte=='Dissolved Ammonia') #create a subset of data for just dissolved ammonia
# 
# unique(df_ammonia$Units) #has both mg/L as N and mg/L
# 
# #dissolved nitrate + nitrite
# df_nitratenitrite <- df_nutrients_comb %>%
#   filter(Analyte=='Dissolved Nitrate + Nitrite') #create a subset of data for just dissolved nitrate + nitrite
# 
# unique(df_nitratenitrite$Units) #has both mg/L as N and mg/L
# 
# #TKN
# df_tkn <- df_nutrients_comb %>%
#   filter(Analyte=='Total Kjeldahl Nitrogen') #create a subset of data for just TKN
# 
# unique(df_tkn$Units) #has both mg/L as N and mg/L
# 
# #total phosphorus
# df_phos <- df_nutrients_comb %>%
#   filter(Analyte=='Total Phosphorus') #create a subset of data for just total phosphorus
# 
# unique(df_phos$Units) #has both mg/L as P and mg/L
# 
# #DON
# df_don <- df_nutrients_comb %>%
#   filter(Analyte=='Dissolved Organic Nitrogen') #create a subset of data for just DON
# 
# unique(df_don$Units) #has both mg/L as N and mg/L
# 
# #dissolved orthophosphate
# df_ortho <- df_nutrients_comb %>%
#   filter(Analyte=='Dissolved ortho-Phosphate') #create a subset of data for just dissolved orthophos
# 
# unique(df_ortho$Units) #has both mg/L as P and mg/L
# 
# #dissolved nitrate
# df_nitrate <- df_nutrients_comb %>%
#   filter(Analyte=='Dissolved Nitrate') #create a subset of data for just dissolved nitrate (O&M only)
# 
# unique(df_nitrate$Units) #only has mg/L

#everything except dissolved nitrate has two different units

#QUESTION: are differences in units for the three analytes due to differences in methods?

#ammonia
# df_ammonia_unit1 <- df_ammonia %>%
#   filter(Units == "mg/L as N")
# 
# unique(df_ammonia_unit1$Method) #only has EPA 350.1
# 
# df_ammonia_unit2 <- df_ammonia %>%
#   filter(Units == "mg/L")
# 
# unique(df_ammonia_unit2$Method) #only has EPA 350.1
# 
# #nitrate + nitrite
# df_nitrate_unit1 <- df_nitratenitrite %>%
#   filter(Units == "mg/L as N")
# 
# unique(df_nitrate_unit1$Method) #has both Std method 4500 and EPA 353.2
# 
# df_nitrate_unit2 <- df_nitratenitrite %>%
#   filter(Units == "mg/L")
# 
# unique(df_nitrate_unit2$Method) #only has EPA 353.2
# 
# #TKN
# df_tkn_unit1 <- df_tkn %>%
#   filter(Units == "mg/L as N")
# 
# unique(df_tkn_unit1$Method) #only has EPA 351.2
# 
# df_tkn_unit2 <- df_tkn %>%
#   filter(Units == "mg/L")
# 
# unique(df_tkn_unit2$Method) #only has EPA 351.2

#ANSWER: differences in units NOT due to differences in methods

# rm(df_ammonia_unit1, df_ammonia_unit2, df_nitrate_unit1, df_nitrate_unit2, df_tkn_unit1, df_tkn_unit2) #remove unit data frames
```

Fix incorrect units

```{r}
# Dave's tip: This can be more succinct with the following code:
df_nutrients_comb <- df_nutrients_comb %>% 
  mutate(
    Units = case_match(
      Analyte,
      "Dissolved Ammonia" ~ "mg/L as N",
      "Dissolved Nitrate" ~ "mg/L as N", # TODO (DB): not sure if this is correct, should it be mg/L as N? - MB: yes, changed to as N
      "Dissolved Nitrate + Nitrite" ~ "mg/L as N",
      "Dissolved Organic Nitrogen" ~ "mg/L as N",
      "Dissolved ortho-Phosphate" ~ "mg/L as P",
      "Total Kjeldahl Nitrogen" ~ "mg/L as N",
      "Total Phosphorus" ~ "mg/L as P"
    )
  )

# #dissolved ammonia 
# #method: EPA 350.1 (should be reported as mg/L as N, based on method)
# df_ammonia_unit <- df_ammonia %>%
#   mutate(Units = recode(Units, "mg/L" = "mg/L as N"))
# 
# unique(df_ammonia_unit$Units) #confirm all say mg/L as N
# 
# #dissolved nitrate + nitrite
# #method: EPA 353.2, EPA 300.0, and Std Method 4500-NO3-, F (should be reported as mg/L as N, based on methods)
# df_nitrate_unit <- df_nitratenitrite %>%
#   mutate(Units = recode(Units, "mg/L" = "mg/L as N"))
# 
# unique(df_nitrate_unit$Units) #confirm all say mg/L as N
# 
# #TKN
# #method: EPA 351.2 (should be reported as mg/L as N, based on method)
# df_tkn_unit <- df_tkn %>%
#   mutate(Units = recode(Units, "mg/L" = "mg/L as N"))
# 
# unique(df_tkn_unit$Units) #confirm all say mg/L as N
# 
# #total phosphorus
# #methods: EPA 365.4 and 365.3 (should all be reported as mg/L as P, based on methods)
# df_phos_unit <- df_phos %>%
#   mutate(Units = recode(Units, "mg/L" = "mg/L as P"))
# 
# unique(df_phos_unit$Units) #confirm all say mg/L as P
# 
# #dissolved orthophosphate
# #methods: EPA 365.1, EPA 300.0, EPA 365.3 (should all be reported as mg/L as P, based on methods)
# df_ortho_unit <- df_ortho %>%
#   mutate(Units = recode(Units, "mg/L" = "mg/L as P"))
# 
# unique(df_ortho_unit$Units) #confirm all say mg/L as P
# 
# #DON
# #method: EPA 351.2/EPA 350.1 (should be reported as mg/L as N, based on method)
# df_don_unit <- df_don %>%
#   mutate(Units = recode(Units, "mg/L" = "mg/L as N"))
# 
# unique(df_don_unit$Units) #confirm all say mg/L as N
# 
# #combine back together
# df_nutrients_comb <- bind_rows(df_nitrate, df_ammonia_unit, df_nitrate_unit, df_tkn_unit, df_phos_unit, df_ortho_unit, df_don_unit)
# 
# rm(df_ammonia, df_nitratenitrite, df_tkn, df_phos, df_don, df_ortho, df_nitrate) #remove analyte data frames
# rm(df_ammonia_unit, df_nitrate_unit, df_tkn_unit, df_phos_unit, df_ortho_unit, df_don_unit) #remove unit data frames
```

### Notes

Combine Description and Notes columns

```{r}
df_nutrients_comb <- df_nutrients_comb %>%
  #combine description and notes columns, remove any NAs
  unite(col = "Notes", Description, Notes, sep = ", ", na.rm = TRUE) %>% 
  # Should probably convert all of the empty strings in Notes to NA
  mutate(Notes = na_if(Notes, ""))
```

### Format Columns

```{r}
df_nutrients_final <- df_nutrients_comb %>%
  rename(Sample_Code = "Sample Code", Reporting_Limit = "Rpt Limit", Station_Number = "Station Number") %>% #clean up column names
  mutate(Date_Time = force_tz(Date_Time,tzone="Etc/GMT+8")) %>% #set time zone to PST
  select(c(Project, Station_Name, Station_Number, Latitude, Longitude, Sample_Code, Date_Time, Analyte, Detection_Condition, Result, Units, Reporting_Limit, Method, Notes)) #reorder columns
  
```

### Plot Raw Data

```{r Plots, fig.height=10, fig.width=10}
quick_plot <- ggplot(df_nutrients_final, aes(x = Date_Time, y = Result, color= Project)) +
  geom_point() + 
  theme_bw() +
  facet_wrap(.~Analyte, ncol = 1)
plot(quick_plot)

#found extreme value of 129 mg/L for dissolved nitrate + nitrite collected by O&M (SWP Routine Monitoring) on 11/18/2014
# df_nutrients_final %>% 
#   arrange(desc(Result)) %>% 
#   View()
```

### Data Checks/Cleaning

Check ranges

```{r}
range(na.omit(df_nutrients_final$Result))

# The min of the results is zero. There shouldn't be any values equal to zero in this dataset. - MB: fixed above in Non-Detects section
# Investigate this further
# df_nutrients_final %>% 
#   filter(Result == 0) %>% 
#   View()
# There are 7 records collected at Lake Oroville that have values equal to zero. They appear to be
# analyzed by Test America and are flagged with "J". Additionally, the 4 TKN records have notes that
# state "Field data only".
# TODO (DB): The 4 TKN records should most likely be removed, and the remaining 3 records may need
# to be converted to Not Detected. Need to investigate this further.

#Morgan: confirmed with data contact, these 7 records were analyzed at Test America, but have a J flag (<RL, >MDL) - these should be non-detects

df_nutrients_clean <- df_nutrients_final %>%
  filter(Result < 50 | is.na(Result)) 

range(df_nutrients_clean$Date_Time)

#some WQA data present for second half of 2024, but all other groups only go through June 2024 due to delay in Bryte reporting - cut off at July

df_nutrients_clean <- df_nutrients_clean %>%
  filter(Date_Time < "2024-07-01 00:00:00 -08") #filter out everything after June 2024

rm(df_nutrients_comb, df_nutrients_final, df_methods)
```

### Plot Clean Data

```{r Plots2, fig.height=12, fig.width=15}
quick_plot <- ggplot(df_nutrients_clean, aes(x = Date_Time, y = Result, color= Project)) +
  geom_point() + 
  theme_bw() +
  facet_wrap(.~Analyte, ncol = 2)
plot(quick_plot)
```

### Check for NA values

```{r}
# df_nutrients_clean %>% 
#   summarize(across(everything(), \(x) sum(is.na(x)))) %>% 
#   pivot_longer(everything()) %>% view()

# Station_Number, Latitude, Longitude, Result, Reporting_Limit, and Notes columns all have some NA
# values. Most of these are expected, but I'm not sure about Station_Number.

# Which records have NA values for Station_Number?
# df_nutrients_clean %>% 
#   filter(is.na(Station_Number)) %>% 
#   count(Project, Station_Name) %>% 
#   arrange(Project, Station_Name) %>% 
#   print(n = Inf)

# Just the LSZ stations from DEMP, and the FRP records have NA values for the Station_Number. 
# TODO (DB): is this okay? MB: yes, it's fine since they are "floating" stations and don't have broad water body numbers
```

### Check duplicate Station/Date/Time/Analyte

```{r}
# TODO: true duplicates, can remove
# While the duplicated function works in this case, it doesn't allow for you to inspect the
# duplicated pairs. I would suggest using the following code to look for duplicated records
df_dups <- df_nutrients_clean %>%
  # filter(duplicated(.))
  add_count(pick(all_of(names(.)))) %>% 
  filter(n > 1) %>% 
  arrange(Project, Sample_Code, Analyte)

#remove all duplicated rows
df_nutrients_clean <- df_nutrients_clean %>% distinct()

unique(df_dups$Project)

rm(df_dups)
```

```{r}
# TODO: key duplicates; are these real?
df_dups1 <- df_nutrients_clean %>%
  add_count(Project, Station_Name, Date_Time, Analyte) %>%
  filter(n > 1) %>% 
  arrange(Project, Station_Name, Date_Time, Analyte)

unique(df_dups1$Project)#FRP, SWP, NCRO-SDTB, WQA, n=2-6

#Morgan: I would think some of these are real, since we can't assume each project only collected one of each sample/analyte per sampling event. It may be helpful to add sample code in there, since that's what FLIMS uses to distinguish "different" samples.

#check with sample code
df_dups2 <- df_nutrients_clean %>%
  add_count(Project, Station_Name, Date_Time, Analyte, Sample_Code) %>%
  filter(n > 1) %>% 
  arrange(Project, Sample_Code, Analyte)

unique(df_dups2$Project) #same four projects as above, but n=2 for all 
range(df_dups2$Date_Time)
```

```{r}
# TODO: this one (without time) has slightly more matches; are they real?
df_dups3 <- df_nutrients_clean %>%
  mutate(Date = as.Date(Date_Time)) %>%
  add_count(Project, Station_Name, Date, Analyte) %>%
  filter(n > 1) %>% 
  arrange(Project, Station_Name, Date, Analyte)

unique(df_dups3$Project) #same four projects

rm(df_dups1, df_dups2, df_dups3)
```

Remove additional lab dups (Lab Dup column only identifies lab dups going back to Dec 2012)

```{r}
# Define grouping variables for each unique duplicate pair
grp_lab_dups <- c("Project", "Station_Name", "Date_Time", "Analyte", "Sample_Code")

# Pull out lab dups and select one at random
df_data_lab_dups <- df_nutrients_clean %>%
  add_count(pick(all_of(grp_lab_dups))) %>%
  filter(n > 1) %>%
  select(-n) %>%
  slice_sample(n = 1, by = all_of(grp_lab_dups))

# Add back the cleaned up lab dup data to the main data frame
df_nutrients_clean <- df_nutrients_clean %>%
  anti_join(
    df_data_lab_dups %>% select(all_of(grp_lab_dups)),
    by = grp_lab_dups
  ) %>%
  bind_rows(df_data_lab_dups)
```

### Round values

```{r}
# Look for Results with large numbers of digits after the decimal place
withr::with_options(
  list(max.print = 2000), 
  print(sort(unique(as.character(df_nutrients_clean$Result))))
)
# There are some values that have 4 digits after the decimal place. Let's look at this more closely.
df_nutrients_dec <- df_nutrients_clean %>% 
  mutate(
    Result_chr = as.character(Result),
    Num_dec_place = str_length(str_remove(Result_chr, "^\\d*\\."))
  )

# Look at which Analytes have results with 4 digits after the decimal place
df_nutrients_dec %>% 
  count(Analyte, Num_dec_place) %>% 
  filter(Num_dec_place > 3)
# Diss Ammonia (n = 2), Diss Nitrate (n = 37), Diss N+N (n = 1), Diss ortho-P (n = 1)

# Look at which Projects have results with 4 digits after the decimal place
df_nutrients_dec %>% 
  count(Project, Num_dec_place) %>% 
  filter(Num_dec_place > 3)
# DEMP (n = 1), O&M (n = 38), WQA (n = 2)

# TODO (DB): Check if we are okay with the results with 4 digits after the decimal place

# TODO: check that this is valid rounding; use case_when if certain nutrients have more sig figs, or
# if its different for different programs
#df_nutrients_clean <- df_nutrients_clean %>% mutate(Result = round(Result, 2),
                              #Reporting_Limit = round(Reporting_Limit, 2))

# Also look at the number of digits after the decimal place for lat-long coordinates - these should
# have about 6 digits
sort(unique(as.character(df_nutrients_clean$Latitude)))
sort(unique(as.character(df_nutrients_clean$Longitude)))
# These all look good

rm(df_nutrients_dec)
```

### Check collection times

```{r}
df_times <- df_nutrients_clean %>% 
  mutate(
    Hour = hour(Date_Time),
    Minute = minute(Date_Time),
    Second = second(Date_Time)
  )

# Check for samples collected at midnight
df_times %>% 
  filter(Hour == 0, Minute == 0, Second == 0) %>% 
  count(Project)
# FRP (n = 1750), O&M (n = 118), WQA (n = 109)
# TODO (DB): These samples most likely did not have a collection time. We should consider adding a
# Date column, and then convert the samples without collection times to NA. This is how we handle
# this situation in discretewq. I think this is probably best because we can be more explicit about
# the records without collection times. Currently someone can misinterpret these samples as being
# collected at midnight.

#convert missing times to NAs and add a date column
df_nutrients_clean <- df_nutrients_clean %>%
  mutate(Date = date(Date_Time)) %>%
  mutate(Hour = hour(Date_Time), Minute = minute(Date_Time), Second = second(Date_Time)) %>%
  mutate(time_sum = Hour + Minute + Second) %>%
  mutate(date_test = case_when((time_sum == 0) ~ NA,  #if the time is 0:00, change it to NA
                    .default = Date_Time)) %>%
  select(-c(Hour, Minute, Second, time_sum, Date_Time)) %>%
  rename(Date_Time = date_test)

# Inspect collection times for oddities
df_times %>% 
  # Remove samples collected at midnight
  filter(!(Hour == 0 & Minute == 0 & Second == 0)) %>% 
  count(Project, Hour) %>% 
  View()
# TODO (DB): DEMP, Tidal Parr, WQA, and YBFMP have some odd collection times (before 5 AM and after
# 5 PM). We may want to inspect these further before publishing the data.
```

### Export as csv file

```{r}
# Arrange data in an more organized way before export
df_nutrients_exp <- df_nutrients_clean %>% arrange(Project, Date_Time, Analyte)

#Reorder columns
df_nutrients_exp <- df_nutrients_exp %>%
  select(c("Project", "Station_Name", "Station_Number", "Latitude", "Longitude", "Sample_Code", "Date", "Date_Time", "Analyte", "Detection_Condition", "Result", "Units", "Reporting_Limit", "Method", "Notes"))

fp_output_discrete <- abs_nutrient_path("../EDI Data Files/nutrient_data_discrete.csv")
df_nutrients_exp %>% 
  # Convert Date_Time to character to avoid having write_csv convert the time zone to UTC
  mutate(Date_Time = format(Date_Time, format = "%Y-%m-%d %H:%M:%S")) %>%
  write_csv(fp_output_discrete)

# Also export to GitHub repo as .rds file (compressed)
df_nutrients_exp %>% saveRDS(here("data/processed/nutrient_data_discrete.rds"))
```

## Continuous Nutrient Data

### 1 - CEMP

Import CEMP SUNA data

```{r}
fp_cemp_hood <- abs_nutrient_path("SRH_SUNA_data.csv")
df_cemp_hood <- read_csv(fp_cemp_hood) #read in Hood data

fp_cemp_vern <- abs_nutrient_path("SJR_SUNA_data.csv")
df_cemp_vern <- read_csv(fp_cemp_vern) #read in Vernalis data

rm(fp_cemp_hood, fp_cemp_vern) #remove file paths
```

Add station names

```{r}
df_cemp_hood <- df_cemp_hood %>%
  add_column(Station_Name = "C3A") #match the corresponding discrete station

df_cemp_vern <- df_cemp_vern %>%
  add_column(Station_Name = "C10A") #match the corresponding discrete station
```

Combine data and format

```{r}
df_cemp_suna <- bind_rows(df_cemp_hood, df_cemp_vern)

df_cemp_suna <- df_cemp_suna %>%
  # This isn't necessary because TIMESTAMP is already a datetime object, just need to rename it
  # mutate(Date_Time = parse_date_time(`TIMESTAMP`, orders = c("Ymd HMS", "Ymd"))) %>% #format date, include both formats since some times missing
  rename(Date_Time = TIMESTAMP, Result = NO3_ave) %>% #rename result column as well
  add_column(Project = "Continuous EMP SUNA Real-Time Monitoring") #add project column

rm(df_cemp_vern, df_cemp_hood)
```

### 2 - WQA

Import WQA RTDF data

```{r}
fp_rtdf_banks <- abs_nutrient_path("RTDF_Banks_NO3.csv")
df_rtdf_banks <- read_csv(fp_rtdf_banks) #read in Banks data

fp_rtdf_gia <- abs_nutrient_path("RTDF_Gianelli_NO3.csv")
df_rtdf_gia <- read_csv(fp_rtdf_gia) #read in Gianelli data

fp_rtdf_jones <- abs_nutrient_path("RTDF_Jones_NO3.csv")
df_rtdf_jones <- read_csv(fp_rtdf_jones) #read in Jones data

fp_rtdf_vern <- abs_nutrient_path("RTDF_Vernalis_NO3.csv")
df_rtdf_vern <- read_csv(fp_rtdf_vern) #read in Vernalis data

rm(fp_rtdf_banks, fp_rtdf_gia, fp_rtdf_jones, fp_rtdf_vern) #remove file paths
```

Add station names

```{r}
df_rtdf_banks <- df_rtdf_banks %>%
  add_column(Station_Name = "Banks Headworks") %>% #match the corresponding discrete station
  select(c(Datetime, NO3_mg_L, Station_Name)) #remove the extra column

df_rtdf_gia <- df_rtdf_gia %>%
  add_column(Station_Name = "Gianelli") #match the corresponding discrete station

df_rtdf_jones <- df_rtdf_jones %>%
  add_column(Station_Name = "Jones Lockup") #match the corresponding discrete station

df_rtdf_vern <- df_rtdf_vern %>%
  add_column(Station_Name = "C10A") #match the corresponding discrete station
```

Combine data and format

```{r}
df_wqa_rtdf <- bind_rows(df_rtdf_banks, df_rtdf_gia, df_rtdf_jones, df_rtdf_vern)

df_wqa_rtdf <- df_wqa_rtdf %>%
  mutate(Date_Time = parse_date_time(Datetime, orders = c("mdY HM", "mdY"))) %>% #format date, include both formats since some times missing
  add_column(Project = "WQA Real-Time Data and Forecasting") %>% #add project column
  rename(Result = NO3_mg_L) %>% #rename result column
  select(-Datetime) #remove old date time column
  
rm(df_rtdf_banks, df_rtdf_gia, df_rtdf_jones, df_rtdf_vern)
```

### Data Integration and Formatting

Combine data from CEMP and WQA and format

```{r}
df_continuous_all <- bind_rows(df_cemp_suna, df_wqa_rtdf)

df_continuous_final <- df_continuous_all %>%
  filter(!(is.na(Result))) %>% #remove NA results (confirmed this is missing data, not non-detect data)
  select(c(Project, Station_Name, Date_Time, Result)) %>% #reorder columns
  # Convert time zone to PST
  mutate(Date_Time = force_tz(Date_Time, tzone = "Etc/GMT+8")) 

rm(df_cemp_suna, df_wqa_rtdf, df_continuous_all)
```

### Plot Raw Data

```{r Plots rtm, fig.height=12, fig.width=12}
quick_plot <- ggplot(df_continuous_final, aes(x = Date_Time, y = Result, color= Project)) +
  geom_point() + 
  theme_bw()+
  facet_wrap(.~Station_Name, ncol = 2)
plot(quick_plot)
```

### Data Checks/Cleaning

```{r}
range(df_continuous_final$Date_Time) #check time frame

range(df_continuous_final$Result) #found some negative values and really high values - confirmed these are not real by data contact

df_continuous_clean <- df_continuous_final %>%
  filter(Result < 60 & Result > 0)
```

### Plot Clean Data

```{r Plots rtm2, fig.height=12, fig.width=12}
quick_plot <- ggplot(df_continuous_clean, aes(x = Date_Time, y = Result, color= Project)) +
  geom_point() + 
  theme_bw()+
  facet_wrap(.~Station_Name, ncol = 2)
plot(quick_plot)
```

### Round values

```{r}
# Look for Results with large numbers of digits after the decimal place
sort(unique(as.character(df_continuous_clean$Result)))

# There are some values that have 4 digits after the decimal place. Let's look at this more closely.
df_continuous_dec <- df_continuous_clean %>% 
  mutate(
    Result_chr = as.character(Result),
    Num_dec_place = str_length(str_remove(Result_chr, "^\\d*\\."))
  )

# Look at which Projects have results with 4 digits after the decimal place
df_continuous_dec %>% 
  count(Project, Num_dec_place) %>% 
  filter(Num_dec_place > 3)
# WQA (n = 11,244)

# TODO (DB): Check if we are okay with the results with 4 digits after the decimal place

# TODO: check that this is valid rounding; use case_when if certain nutrients have more sig figs, or if its dif for dif programs

# I didn't run it because I know this is dif for dif continuous ones.

# df_continuous_clean <- df_continuous_clean %>% mutate(Result = round(Result, 2),
#                               Reporting_Limit = round(Reporting_Limit, 2))

rm(df_continuous_dec)
```

### Check duplicate Station/Date/Time/Analyte

```{r}
# none are fully duplicates
df_dup_test <- df_continuous_clean %>%
  # filter(duplicated(.))
  add_count(pick(all_of(names(.)))) %>% 
  filter(n > 1) %>% 
  arrange(Project, Station_Name, Date_Time)

rm(df_dup_test)
```

```{r}
# TODO: some have duplicate "keys"; are these real?
dup_check1 <- df_continuous_clean %>%
  add_count(Station_Name, Date_Time) %>%
  filter(n > 1) %>% 
  arrange(Date_Time, Station_Name)

unique(dup_check1$Station_Name)

#only station that has dups is C10A, which is present in both the CEMP SUNA data and the WQA RTDF data - could just be overlap between projects

dup_check2 <- df_continuous_clean %>%
  add_count(Project, Station_Name, Date_Time) %>%
  filter(n > 1)

#confirmed, was just overlap between the two projects (no dups when project is included)

rm(dup_check1, dup_check2)
```

### Export as csv file

```{r}
# Arrange data in an more organized way before export
df_continuous_exp <- df_continuous_clean %>% arrange(Project, Station_Name, Date_Time)

fp_output_continuous <- abs_nutrient_path("../EDI Data Files/nutrient_data_continuous.csv")

# Use write_excel_csv instead because write_csv converts the time zone to UTC
df_continuous_exp %>% 
  # Convert Date_Time to character to avoid having write_csv convert the time zone to UTC
  mutate(Date_Time = format(Date_Time, format = "%Y-%m-%d %H:%M:%S")) %>% 
  write_csv(fp_output_continuous)

# Also export to GitHub repo as .rds file (compressed)
df_continuous_exp %>% saveRDS(here("data/processed/nutrient_data_continuous.rds"))
```

