---
title: "Process Input Data for Models"
author: "Dave Bosworth"
date: today
date-format: long
format:
  html:
    toc: true
    toc-expand: 2
    code-fold: show
    code-tools: true
execute: 
  message: false
  warning: false
---

# Purpose

Prepare input data for models used in the DWR nutrient synthesis project.

# Global Code

```{r load packages}
# Load packages
library(tidyverse)
library(rlang)
library(glue)
library(hms)
library(scales)
# Make sure we are using `deltamapr` version 1.0.1, commit fe34697b3d1aaa2945bbfc647582a19e251abf67
# install.packages("devtools")
# devtools::install_github("InteragencyEcologicalProgram/deltamapr", ref = "fe34697b3d1aaa2945bbfc647582a19e251abf67")
library(deltamapr)
# Make sure we are using `discretewq` version 2.4.0.9000, commit b68425c2eca88015b5e01ae6029f84b2a3af2ea8
# devtools::install_github("InteragencyEcologicalProgram/discretewq", ref = "b68425c2eca88015b5e01ae6029f84b2a3af2ea8")
library(discretewq)
library(sf)
library(mapview)
library(EDIutils)
library(fs)
library(here)
library(conflicted)

# Declare package conflict preferences
conflicts_prefer(dplyr::filter(), hms::hms())
```

```{r session info}
# Run session info to display package versions
devtools::session_info()
```

Create functions used in this document:

```{r functions}
# Get data entity names for specified EDI ID
get_edi_data_entities <- function(edi_id) {
  df_data_ent <- read_data_entity_names(edi_id)
  inform(c(
    "i" = paste0(
      "Data entities for ", edi_id, " include:\n",
      paste(df_data_ent$entityName, collapse = "\n"), "\n"
    ))
  )
  return(df_data_ent$entityName)
}

# Download specified data entities from an EDI package and save raw bytes files to a temporary
# directory
get_edi_data <- function(edi_id, entity_names) {
  df_data_ent <- read_data_entity_names(edi_id)
  df_data_ent_filt <- df_data_ent %>% filter(entityName %in% entity_names)
  
  ls_data_raw <-
    map(df_data_ent_filt$entityId, \(x) read_data_entity(edi_id, entityId = x)) %>%
    set_names(df_data_ent_filt$entityName)
  
  temp_dir <- tempdir()
  for (i in 1:length(ls_data_raw)) {
    file_raw <- file.path(temp_dir, glue("{names(ls_data_raw)[i]}.bin"))
    con <- file(file_raw, "wb")
    writeBin(ls_data_raw[[i]], con)
    close(con)
  }
}

# Replace values below the reporting limit with simulated values between `min_val` and the RL
replace_blw_rl <- function(df, min_val = 0, seed = 1) {
  # Pull out values that are below the RL
  df_blw_rl <- df %>% filter(Sign == "<")
  
  # Replace below RL values with simulated ones
  withr::with_seed(
    # Set seed for reproducibility
    seed = seed,
    df_blw_rl_sim <- df_blw_rl %>% 
      mutate(Result = round(runif(nrow(df_blw_rl), min = min_val, max = Result), 6))
  )
  
  # Add simulated values back to main data frame
  df %>% filter(Sign != "<") %>% bind_rows(df_blw_rl_sim)
}

# Add variable for season as a factor
add_season <- function(df) {
  df %>% mutate(
    Season = case_match(
      Month,
      3:5 ~ "Spring",
      6:8 ~ "Summer",
      9:11 ~ "Fall",
      c(12, 1, 2) ~ "Winter"
    ),
    Season = factor(Season, levels = c("Winter", "Spring", "Summer", "Fall"))
  )
}
```

Load globally-used data:

```{r load global data}
# Import region assignments
df_regions <- read_csv(here("data/raw/region_assignments.csv"))

# Load Delta EDSM shapefile and only keep SubRegions east of Carquinez Straight
sf_delta <- R_EDSM_Subregions_Mahardja_FLOAT %>% 
  filter(
    !SubRegion %in% c(
      "Carquinez Strait", 
      "Lower Napa River", 
      "San Francisco Bay",
      "San Pablo Bay",
      "South Bay",
      "Upper Napa River" 
    )
  ) %>% 
  select(SubRegion) %>% 
  # Add region assignments to the SubRegion shapefile
  left_join(df_regions, by = join_by(SubRegion)) %>% 
  relocate(Region, .after = SubRegion)
```

# Regions Map

Create map showing SubRegions and Regions used to process the data.

```{r}
mapview(sf_delta, zcol = "Region", alpha.regions = 0.5)
```

# DWR Integrated Discrete Nutrient Data

## Import Data

```{r dwr nutr import data}
# Import nutrient data set
df_nutr <- readRDS(here("data/processed/nutrient_data_discrete.rds"))

# Import phytoplankton data from FRP's EDI publication to assign lat-long coordinates to FRP's
  # nutrient data in the DWR integrated discrete nutrient data set
edi_data_ent <- get_edi_data_entities("edi.269.5")
edi_data_ent_sub <- str_subset(edi_data_ent, "^phyto")
get_edi_data("edi.269.5", edi_data_ent_sub)
temp_files <- list.files(tempdir(), full.names = TRUE)

df_frp_phyto_coord <- read_csv(str_subset(temp_files, edi_data_ent_sub)) %>% 
  drop_na(LatitudeStart, LongitudeStart) %>% 
  summarize(
    Latitude = mean(LatitudeStart),
    Longitude = mean(LongitudeStart),
    .by = Location
  )
```

## Prepare Data

```{r dwr nutr prepare data}
# Prepare data for aggregation
df_nutr_c1 <- df_nutr %>% 
  # Add lat-long coordinates for FRP stations from the phytoplankton data set
  filter(Project == "Fish Restoration Program") %>% 
  select(-c(Latitude, Longitude)) %>% 
  left_join(df_frp_phyto_coord, by = join_by(Station_Name == Location)) %>% 
  bind_rows(df_nutr %>% filter(Project != "Fish Restoration Program")) %>% 
  # Remove records without lat-long coordinates
  drop_na(Latitude, Longitude) %>% 
  # Assign SubRegions to the stations
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326) %>%
  st_transform(crs = st_crs(sf_delta)) %>%
  st_join(sf_delta, join = st_intersects) %>%
  # Remove any data outside our subregions of interest
  filter(!is.na(SubRegion)) %>%
  st_drop_geometry()
```

Remove various duplicated records so that there is only one sample per station-day.

```{r dwr nutr rm dt dups}
# Clean up duplicated records that share same Date_Time - it might be more appropriate to address
  # this in the nutrient data integration code. TODO: talk to Morgan about this
df_nutr_dt_dups <- df_nutr_c1 %>% 
  add_count(Project, Station_Name, Date_Time, Analyte) %>% 
  filter(n > 1) %>% 
  select(-n) %>% 
  drop_na(Date_Time)

# Keep the records with the "earliest" Sample_Code
df_nutr_dt_dups_fixed <- df_nutr_dt_dups %>% 
  arrange(Project, Station_Name, Date_Time, Analyte, Sample_Code) %>% 
  distinct(Project, Station_Name, Date_Time, Analyte, .keep_all = TRUE)

df_nutr_c2 <- df_nutr_c1 %>% 
  anti_join(df_nutr_dt_dups) %>% 
  bind_rows(df_nutr_dt_dups_fixed)
```

```{r dwr nutr rm dups no dt}
# Clean up duplicated records that share same Date and don't have a collection time
df_nutr_dups_nodt <- df_nutr_c2 %>% 
  add_count(Project, Station_Name, Date, Analyte) %>% 
  filter(n > 1, is.na(Date_Time)) %>% 
  select(-n)

# Keep the records with the "earliest" Sample_Code
df_nutr_dups_nodt_fixed <- df_nutr_dups_nodt %>% 
  arrange(Project, Station_Name, Date, Analyte, Sample_Code) %>% 
  distinct(Project, Station_Name, Date, Analyte, .keep_all = TRUE)

df_nutr_c3 <- df_nutr_c2 %>% 
  anti_join(df_nutr_dups_nodt) %>% 
  bind_rows(df_nutr_dups_nodt_fixed)
```

```{r dwr nutr rm dups same day}
# Filter data so that there is only one sample per station-day by choosing the data point closest 
  # to noon
df_nutr_daily_dups <- df_nutr_c3 %>% 
  add_count(Project, Station_Name, Date, Analyte) %>% 
  filter(n > 1) %>% 
  select(-n)

df_nutr_daily_dups_fixed <- df_nutr_daily_dups %>% 
  # Create variable for time and calculate difference from noon for each data point
  mutate(
    Time = as_hms(Date_Time),
    Noon_diff = abs(hms(hours = 12) - Time)
  ) %>%
  group_by(Project, Station_Name, Date, Analyte) %>%
  # Select only 1 data point per Station, Date, and Analyte, choose data closest to noon
  filter(Noon_diff == min(Noon_diff)) %>%
  # When points are equidistant from noon, select earlier point
  filter(Time == min(Time)) %>%
  ungroup() %>% 
  select(-c(Time, Noon_diff))

df_nutr_c4 <- df_nutr_c3 %>% 
  anti_join(df_nutr_daily_dups) %>% 
  bind_rows(df_nutr_daily_dups_fixed)
```

Replace values below the reporting limit with simulated values.

```{r dwr nutr replace blw rl vals}
# Replace values below the reporting limit with simulated values
df_nutr_c5 <- df_nutr_c4 %>% 
  mutate(
    Sign = if_else(Detection_Condition == "Not Detected", "<", "="),
    Result = if_else(Detection_Condition == "Not Detected", Reporting_Limit, Result),
    .keep = "unused"
  ) %>% 
  # Omit <RL values with reporting limits that are greater than the 95th percentile of the values
    # for the parameter
  mutate(q95 = quantile(Result, probs = 0.95), .by = Analyte) %>% 
  filter(!(Sign == "<" & Result > q95)) %>% 
  select(-q95) %>% 
  replace_blw_rl()
```

## Aggregate Data

```{r dwr nutr calc monthly values}
# Calculate Monthly-Regional averages
df_nutr_avg_mo <- df_nutr_c5 %>% 
  mutate(
    Year = year(Date),
    Month = month(Date)
  ) %>% 
  summarize(
    Result = mean(Result),
    Num_samples = n(),
    .by = c(Year, Month, Region, Analyte)
  )
```

```{r dwr nutr calc seasonal values}
# Further calculate Seasonal-Regional averages using Adjusted calendar year: December-November, with
  # December of the previous calendar year included with the following year
df_nutr_avg_seas <- df_nutr_avg_mo %>% 
  mutate(YearAdj = if_else(Month == 12, Year + 1, Year)) %>% 
  add_season() %>% 
  summarize(
    Result = mean(Result),
    Num_samples = sum(Num_samples),
    .by = c(YearAdj, Season, Region, Analyte)
  )
```

```{r dwr nutr conv monthly seasonal to wide}
# Convert both sets of averages to wide format
ls_nutr_avg_wide <- lst(df_nutr_avg_mo, df_nutr_avg_seas) %>% 
  map(
    \(x) mutate(x,
      Analyte = case_match(
        Analyte,
        "Dissolved Ammonia" ~ "DissAmmonia",
        "Dissolved Nitrate" ~ "DissNitrate",
        "Dissolved Nitrate + Nitrite" ~ "DissNitrateNitrite",
        "Dissolved Organic Nitrogen" ~ "DON",
        "Dissolved ortho-Phosphate" ~ "DissOrthophos",
        "Total Kjeldahl Nitrogen" ~ "TKN",
        "Total Phosphorus" ~ "TotPhos"
      )
    ) %>% 
    arrange(Analyte, .locale = "en") %>% 
    select(-Num_samples) %>% 
    pivot_wider(names_from = Analyte, values_from = Result)
  )
```

## Sampling Effort Plots

### Monthly Values

```{r dwr nutr samp effort monthly values}
ls_nutr_avg_mo_plt <- df_nutr_avg_mo %>% 
  mutate(Month = month(Month, label = TRUE)) %>% 
  complete(nesting(Year, Month), Region, Analyte) %>%
  nest(.by = Analyte) %>% 
  deframe() %>% 
  imap(
    \(x, idx) ggplot(data = x, aes(x = Year, y = Region, fill = Num_samples)) +
      geom_tile() +
      facet_wrap(vars(Month)) +
      ggtitle(idx) +
      scale_y_discrete(drop = FALSE) +
      scale_x_continuous(
        breaks = breaks_pretty(10), 
        expand = expansion(mult = 0.02)
      ) +
      scale_fill_viridis_c(name = "Number of\nSamples", na.value = "transparent") +
      theme_bw() +
      theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
  )
```

::: {.panel-tabset}

```{r dwr nutr print samp effort monthly values}
#| echo: false
#| results: asis
#| fig-width: 8
#| fig-height: 6

for (i in 1:length(ls_nutr_avg_mo_plt)) {
  # Create subheadings for each Parameter
  cat("#### ", names(ls_nutr_avg_mo_plt[i]), "\n\n")
  # Print plot
  print(ls_nutr_avg_mo_plt[[i]])
  cat("\n\n")
}
```

:::

### Seasonal Values

```{r dwr nutr samp effort seasonal values}
#| fig-width: 8
#| fig-height: 9

df_nutr_avg_seas %>% ggplot(aes(x = YearAdj, y = Region, fill = Num_samples)) +
  geom_tile() +
  facet_grid(
    rows = vars(Analyte), 
    cols = vars(Season), 
    labeller = label_wrap_gen(width = 20)
  ) +
  scale_x_continuous(
    breaks = breaks_pretty(10), 
    expand = expansion(mult = 0.02)
  ) +
  scale_fill_viridis_c(name = "Number of\nSamples") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```


# Integrated Discrete WQ Data from discretewq

## Import Data

```{r discretewq import data}
# Import full integrated data set from discretewq
df_dwq <- wq(
  Sources = c(
    "20mm", "Baystudy", "DJFMP", "DOP", "EDSM", "EMP", "FMWT", "NCRO", "SDO", "SKT", "SLS", "STN",
    "Suisun", "USBR", "USGS_CAWSC", "USGS_SFBS", "YBFMP"
  )
)
```

## Prepare Data

```{r discretewq prepare data}
# Define water quality parameters to keep from discretewq
dwq_param <- c(
  "Microcystis", "Temperature", "Salinity", "DissolvedOxygen", "pH", "TurbidityNTU",
  "TurbidityFNU", "Chlorophyll"
)

# Prepare data for aggregation
df_dwq_c1 <- df_dwq %>% 
  select(
    Source, Station, Latitude, Longitude, Date, Datetime, Year, Month, 
    all_of(dwq_param), Chlorophyll_Sign
  ) %>% 
  # Convert Datetime to PST
  mutate(Datetime = with_tz(Datetime, tzone = "Etc/GMT+8")) %>%
  # Remove rows where all parameters are NA
  filter(!if_all(all_of(dwq_param), is.na)) %>% 
  # Remove records without lat-long coordinates
  drop_na(Latitude, Longitude) %>% 
  # Assign SubRegions to the stations
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326) %>%
  st_transform(crs = st_crs(sf_delta)) %>%
  st_join(sf_delta, join = st_intersects) %>%
  # Remove any data outside our subregions of interest
  filter(!is.na(SubRegion)) %>%
  st_drop_geometry() %>% 
  # Remove data prior to 2009 and after 2024
  filter(Year %in% 2009:2024) %>% 
  # Only keep the stations located in Suisun, Montezuma, and Nurse Sloughs from the Suisun Marsh
    # survey
  filter(!(Source == "Suisun" & !str_detect(Station, "^SU|^MZ|^NS")))

# Pivot the parameter columns to long data structure
df_dwq_c2 <- df_dwq_c1 %>% 
  # Add a value suffix to the parameter columns
  rename_with(\(x) paste0(x, "_Result"), all_of(dwq_param)) %>% 
  # Pivot parameters longer, creating a sign and result column for each parameter
  pivot_longer(
    cols = ends_with(c("Sign", "Result")),
    names_to = c("Parameter", ".value"),
    names_pattern = "(.*)_(.*)"
  ) %>% 
  drop_na(Result) %>% 
  replace_na(list(Sign = "="))
```

Consolidate Turbidity parameters and prefer TurbidityNTU when both NTU and FNU were collected.

```{r discretewq consolidate turbidity}
df_dwq_c3 <- df_dwq_c2 %>% 
  mutate(Parameter2 = if_else(str_detect(Parameter, "^Turbidity"), "Turbidity", Parameter))

df_dwq_turb_dups <- df_dwq_c3 %>% 
  filter(Parameter2 == "Turbidity") %>% 
  add_count(Source, Station, Datetime) %>% 
  filter(n > 1) %>% 
  select(-n)

df_dwq_turb_dups_fixed <- df_dwq_turb_dups %>% filter(Parameter == "TurbidityNTU")

df_dwq_c4 <- df_dwq_c3 %>% 
  anti_join(df_dwq_turb_dups) %>% 
  bind_rows(df_dwq_turb_dups_fixed) %>% 
  select(-Parameter) %>% 
  rename(Parameter = Parameter2)
```

Remove various duplicated records so that there is only one sample per station-day.

```{r discretewq rm dt dups}
# Clean up the duplicated records that share same Datetime from EDSM
df_dwq_dt_dups <- df_dwq_c4 %>% 
  add_count(Source, Station, Datetime, Parameter) %>% 
  filter(n > 1) %>% 
  select(-n)

df_dwq_dt_dups_fixed <- df_dwq_dt_dups %>% 
  distinct(Source, Station, Datetime, Parameter, .keep_all = TRUE)

df_dwq_c5 <- df_dwq_c4 %>% 
  anti_join(df_dwq_dt_dups) %>% 
  bind_rows(df_dwq_dt_dups_fixed)
```

```{r discretewq rm dups same day}
# Filter data so that there is only one sample per station-day by choosing the data point closest 
  # to noon
df_dwq_daily_dups <- df_dwq_c5 %>% 
  add_count(Source, Station, Date, Parameter) %>% 
  filter(n > 1) %>% 
  select(-n)

df_dwq_daily_dups_fixed <- df_dwq_daily_dups %>% 
  # Create variable for time and calculate difference from noon for each data point
  mutate(
    Time = as_hms(Datetime),
    Noon_diff = abs(hms(hours = 12) - Time)
  ) %>%
  group_by(Source, Station, Date, Parameter) %>%
  # Select only 1 data point per Station, Date, and Parameter, choose data closest to noon
  filter(Noon_diff == min(Noon_diff)) %>%
  # When points are equidistant from noon, select earlier point
  filter(Time == min(Time)) %>%
  ungroup() %>% 
  select(-c(Time, Noon_diff))

df_dwq_c6 <- df_dwq_c5 %>% 
  anti_join(df_dwq_daily_dups) %>% 
  bind_rows(df_dwq_daily_dups_fixed)
```

Remove values that are out of range of reasonable limits for the parameter

```{r discretewq rm outliers}
df_dwq_c7 <- df_dwq_c6 %>% 
  # Temperature value of 2.9 collected on 8/12/2014 in Suisun Marsh
  filter(!(Parameter == "Temperature" & Result < 3)) %>% 
  # Temperature value of 116 collected on 3/23/2017 at USGS-11447650
  filter(!(Parameter == "Temperature" & Result > 35)) %>% 
  # DO values greater than 20 (these are obviously % Saturation)
  filter(!(Parameter == "DissolvedOxygen" & Result > 20)) %>% 
  # pH value of 2.56 collected on 7/2/2014 in the Yolo Bypass
  filter(!(Parameter == "pH" & Result < 3)) %>% 
  # Two Turbidity values greater than 1000 collected in early Sept 2017 in the South Delta
  filter(!(Parameter == "Turbidity" & Result > 1000)) %>% 
  # A few Chlorophyll values less than or equal to zero
  filter(!(Parameter == "Chlorophyll" & Result <= 0))
```

## Aggregate Data

```{r discretewq calc monthly values}
# Calculate Monthly-Regional averages
df_dwq_avg_mo <- df_dwq_c7 %>% 
  # Convert pH values to H+ concentration before averaging
  mutate(Result = if_else(Parameter == "pH", 10^-Result, Result)) %>% 
  # Replace Chlorophyll values below the reporting limit with simulated values
  replace_blw_rl() %>% 
  summarize(
    Result = mean(Result),
    Num_samples = n(),
    .by = c(Year, Month, Region, Parameter)
  )
```

```{r discretewq calc seasonal values}
# Further calculate Seasonal-Regional averages using Adjusted calendar year: December-November, with
  # December of the previous calendar year included with the following year
df_dwq_avg_seas <- df_dwq_avg_mo %>% 
  mutate(YearAdj = if_else(Month == 12, Year + 1, Year)) %>% 
  add_season() %>% 
  summarize(
    Result = mean(Result),
    Num_samples = sum(Num_samples),
    .by = c(YearAdj, Season, Region, Parameter)
  )
```

```{r discretewq conv monthly seasonal to wide}
# Convert both sets of averages to wide format
ls_dwq_avg_wide <- lst(df_dwq_avg_mo, df_dwq_avg_seas) %>% 
  map(
    \(x) select(x, -Num_samples) %>% 
      pivot_wider(names_from = Parameter, values_from = Result) %>% 
      mutate(
        # Round Microcystis to nearest whole number
        Microcystis = round(Microcystis),
        # Convert average H+ concentration to average pH
        pH = -log10(pH)
      )
  )
```

## Sampling Effort Plots

### Monthly Values

```{r discretewq samp effort monthly values}
ls_dwq_avg_mo_plt <- df_dwq_avg_mo %>% 
  filter(Year %in% 2010:2024) %>% 
  mutate(Month = month(Month, label = TRUE)) %>% 
  complete(nesting(Year, Month), Region, Parameter) %>%
  nest(.by = Parameter) %>% 
  deframe() %>% 
  imap(
    \(x, idx) ggplot(data = x, aes(x = Year, y = Region, fill = Num_samples)) +
      geom_tile() +
      facet_wrap(vars(Month)) +
      ggtitle(idx) +
      scale_y_discrete(drop = FALSE) +
      scale_x_continuous(
        breaks = breaks_pretty(10), 
        expand = expansion(mult = 0.02)
      ) +
      scale_fill_viridis_c(name = "Number of\nSamples", na.value = "transparent") +
      theme_bw() +
      theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
  )
```

::: {.panel-tabset}

```{r discretewq print samp effort monthly values}
#| echo: false
#| results: asis
#| fig-width: 8
#| fig-height: 6

for (i in 1:length(ls_dwq_avg_mo_plt)) {
  # Create subheadings for each Parameter
  cat("#### ", names(ls_dwq_avg_mo_plt[i]), "\n\n")
  # Print plot
  print(ls_dwq_avg_mo_plt[[i]])
  cat("\n\n")
}
```

:::

### Seasonal Values

```{r discretewq samp effort seasonal values}
#| fig-width: 8
#| fig-height: 9

df_dwq_avg_seas %>% 
  filter(YearAdj %in% 2010:2024) %>% 
  ggplot(aes(x = YearAdj, y = Region, fill = Num_samples)) +
    geom_tile() +
    facet_grid(rows = vars(Parameter), cols = vars(Season)) +
    scale_x_continuous(
      breaks = breaks_pretty(10), 
      expand = expansion(mult = 0.02)
    ) +
    scale_fill_viridis_c(name = "Number of\nSamples") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```


# DAYFLOW

## Download Data

Download data from CNRA data portal: <https://data.cnra.ca.gov/dataset/dayflow>

This code chunk is set to not evaluate upon rendering and is only here for transparency sake. Set eval to `TRUE` to run the download data code if desired.

```{r dayflow download data}
#| eval: false

# Download data from URL
dayflow_1997_2023 <- read_csv(
  "https://data.cnra.ca.gov/dataset/06ee2016-b138-47d7-9e85-f46fae674536/resource/21c377fe-53b8-4bd6-9e1f-2025221be095/download/dayflow-results-1997-2023.csv"
)
dayflow_2024 <- read_csv(
  "https://data.cnra.ca.gov/dataset/06ee2016-b138-47d7-9e85-f46fae674536/resource/6a7cb172-fb16-480d-9f4f-0322548fee83/download/dayflowcalculations2024.csv"
)

# Save local copies of data
lst(dayflow_1997_2023, dayflow_2024) %>%
  iwalk(\(x, idx) write_csv(x, file = here("data/raw", paste0(idx, ".csv"))))

rm(dayflow_1997_2023, dayflow_2024)
```

## Import Data

```{r dayflow import data}
# Define file paths of DAYFLOW data
fp_dayflow <- dir_ls(here("data/raw"), regexp = "dayflow.+\\.csv$")

# Import data
ls_dayflow <- fp_dayflow %>% 
  map(read_csv) %>% 
  set_names(\(x) str_extract(x, "(?<=raw/)dayflow.+(?=\\.csv)"))
```

## Prepare Data

```{r dayflow prepare data}
# Clean up and combine data
df_dayflow <- ls_dayflow %>% 
  map_at("dayflow_1997_2023", \(x) mutate(x, Date = mdy(Date))) %>% 
  map(\(x) select(x, Date, SAC, SJR, TOT)) %>% 
  list_rbind() %>% 
  mutate(Year = year(Date), Month = month(Date)) %>% 
  # Remove data prior to 2009 and after 2024
  filter(Year %in% 2009:2024)
```

## Aggregate Data

```{r dayflow calc monthly values}
# Calculate monthly totals
df_dayflow_tot_mo <- df_dayflow %>% 
  summarize(
    Sac_Inflow = sum(SAC),
    SJR_Inflow = sum(SJR),
    Total_Inflow = sum(TOT),
    .by = c(Year, Month)
  )
```

```{r dayflow calc seasonal values}
# Further calculate seasonal totals using Adjusted calendar year: December-November, with December
  # of the previous calendar year included with the following year
df_dayflow_tot_seas <- df_dayflow_tot_mo %>% 
  mutate(YearAdj = if_else(Month == 12, Year + 1, Year)) %>% 
  add_season() %>% 
  summarize(across(ends_with("Inflow"), sum), .by = c(YearAdj, Season))
```


# Combine All Data

```{r combine monthly values}
# Create data frame that contains all possible combinations of year, month, and region
df_yr_mo_reg <- 
  expand_grid(
    Year = 2010:2024,
    Month = 1:12,
    Region = unique(df_regions$Region)
  ) %>% 
  # Remove rows after June 2024
  filter(!(Year == 2024 & Month > 6)) %>% 
  # Create Month-Year column
  mutate(MonthYear = paste(month(Month, label = TRUE), Year, sep = "-"), .before = Year)

# Combine all Monthly-Regional averages (or monthly totals) into one data frame
monthly_values <- 
  list(
    df_yr_mo_reg, 
    ls_nutr_avg_wide$df_nutr_avg_mo, 
    ls_dwq_avg_wide$df_dwq_avg_mo,
    df_dayflow_tot_mo
  ) %>% 
  reduce(left_join)
```

```{r combine seasonal values}
# Create data frame that contains all possible combinations of year, season, and region
df_yr_seas_reg <- df_yr_mo_reg %>% 
  add_season() %>% 
  distinct(YearAdj = Year, Season, Region) %>% 
  # Remove Summer 2024 because of low sampling
  filter(!(YearAdj == 2024 & Season == "Summer"))

# Combine all Seasonal-Regional averages (or seasonal totals) into one data frame
seasonal_values <- 
  list(
    df_yr_seas_reg, 
    ls_nutr_avg_wide$df_nutr_avg_seas, 
    ls_dwq_avg_wide$df_dwq_avg_seas,
    df_dayflow_tot_seas
  ) %>% 
  reduce(left_join)
```


# Export Processed Data

Export the monthly and seasonal data both as .csv and .rds files for the models. Also export the Delta SubRegions and Regions shapefile.

```{r export final data}
ls_final_data <- lst(monthly_values, seasonal_values)
fp_processed <- here("data/processed")

# Export data as csv files
ls_final_data %>% 
  iwalk(\(x, idx) write_csv(x, file = paste0(fp_processed, "/", idx, ".csv")))

# Export data as rds files
ls_final_data %>% 
  iwalk(\(x, idx) saveRDS(x, file = paste0(fp_processed, "/", idx, ".rds")))

# Export Delta SubRegions/Regions shapefile
sf_delta %>% write_sf(here("data/spatial/delta_subregions.shp"))
```

