# Analysis Example

Quick example showing the steps needed for analyzing data for SEM.

```{r}
library(tidyverse)
library(emmeans)
library(car)
library(lme4)
library(lmerTest)
library(lubridate)
theme_set(theme_bw())
options(contrasts = c('contr.sum', 'contr.poly'))
source('code/functions/ns_funcs.R')
```

# Read in data

```{r}
fp_monthly <- abs_ns_path('Data/monthly_values.csv')
df_monthly <- read_csv(fp_monthly) %>%
  mutate(Region = factor(Region),
         Date = make_date(Year, Month, 1)) %>%
  arrange(Region, Date)

fp_seasonal <- abs_ns_path('Data/seasonal_values.csv')
df_seasonal <- read_csv(fp_seasonal) %>%
  mutate(Region = factor(Region))
```

# Exploratory analysis (monthly)

## Balance check

How many samples do we have per region?

```{r}
df_monthly %>% 
  count(Region)
```

Pretty even, but how many are non-NAs for relevant variables?

```{r}
df_monthly %>%
  group_by(Region) %>%
  summarize(
    across(
      DissAmmonia:Total_Inflow,
      ~ sum(!is.na(.)),
      .names = '{col}'
    ),
    .groups = 'drop'
  )
```

There are missing values for certain analytes. The categories are:

**balanced**: Temperature, Salinity, Turbidity, DissolvedOxygen, Inflows

**relatively balanced (\>= 75%)**: Chlorophyll, pH, MVI

**missing data (Suisun Marsh)**: DissAmmonia, DissNitrateNitrite, DissOrthophos, DON, TKN, TotPhos

**missing a lot of data**: DissNitrate

We can break this down by month to see if the missing data is interpolable or not:

```{r}
df_missing <- df_monthly %>%
  pivot_longer(
    cols = DissAmmonia:Total_Inflow,
    names_to = 'Variable',
    values_to = 'Value'
  ) %>%
  mutate(
    Missing = factor(
      if_else(is.na(Value), 1, 0),
      levels = c(0,1),
      labels = c('Present','Missing')
    ),
    Month = floor_date(Date, 'month')
  )

plots <- df_missing %>%
  group_by(Variable) %>%
  group_split() %>%
  lapply(function(df) {
    ggplot(df, aes(x = Month, y = Region, fill = Missing)) +
      geom_tile(color = 'grey20', linewidth = 0.3, width = 35) +
      scale_fill_manual(
        values = c('Present' = 'lightblue', 'Missing' = 'white')
      ) +
      scale_x_date(
        date_breaks = '6 months',
        date_labels = '%m-%Y'
      ) +
      theme_minimal() +
      labs(
        title = paste0('Missing Data: ', unique(df$Variable)),
        fill = 'P/A'
      ) +
      theme(
        axis.text.x = element_text(angle = 90, hjust = 1, size = 6),
        panel.grid = element_blank()
      )
  })

plots
```

Findings:

**No data until 02-2017 for Suisun Marsh:** DON, DissAmmonia, DissNitrateNitrite, DissOrthophos, TKN, TotPhos

-   rest of missing data is interpolable

**Lots of missing data:** DissNitrate

**Large gaps until 06-2015:** MVI

-   all 5 month gaps between Jan-May

**Missing after 01-2023 for Suisun Marsh/Suisun Bay:** pH

**Small interpolable gaps:** Chlorophyll

**No gaps:** Temperature, Salinity, Turbidity, DissolvedOxygen, Inflows

## Discussion

A few analytes can be used for the entire 01-2010 through 06-2024 time range:

-   **Fully ready:** Temperature, Salinity, Turbidity, DissolvedOxygen, Inflows

-   **Needs minor interpolation:** Chlorophyll

A few could be used for the entire range (with minor interpolation) for all regions except Suisun Marsh. The options are:

-   cut the analysis range to begin on 02-2017

-   remove Suisun Marsh from primary analyses, keep full range

    -   could run separately on narrower time frame

-   analyze with full dataset, interpret with unbalanced design in mind

pH is missing data after 01-2023 for two regions; this feels like an error to me, so we might want to double check the original data. If not, the options are similar to above:

-   cut the analysis range to end at 01-2023

-   remove Suisun Marsh/Suisun Bay from primary analyses

-   analyze with full dataset, interpret with unbalanced design in mind

MVI has multiple data gaps and some large, non-interpolable gaps. Options are similar to above. We also have chlorophyll data as a response variable to ground truth our results if analyses look off with MVI.

DissNitrate is missing large amounts of data and should be excluded.

# Data Processing

First I remove columns I won't be using for simplicity.

```{r}
df_monthly_rm <- df_monthly %>%
  select(-c(DissNitrate, DON, Microcystis, SJR_Inflow, Sac_Inflow, TKN, TotPhos))
```

## Note Data Shape

First I look at the shape of the variables to get a sense of what the data looks like.

```{r}
df_plt <- df_monthly_rm %>%
  pivot_longer(
    cols = DissAmmonia:Total_Inflow,
    names_to = 'Analyte',
    values_to = 'Value'
  )

ggplot(df_plt, aes(x = Value)) +
  geom_histogram(
    color = 'black',
    fill = 'lightgray'
  ) +
  facet_wrap(~Analyte, scales = 'free')
```

They're pretty much all appear right-skewed to various degrees except for temperature, which almost appears bivariate, and pH which is slightly left skewed. A few, like chlorophyll and pH, indicate outliers exist.

## Note Outliers

I next note extreme outliers. Even if they're real, they can influence regressions either by influencing the formula or inflating the error.

If, later on (eg. after model diagnostics), I decide to remove any of these, this is where I'd do it.

```{r}
df_plt <- df_plt %>%
  group_by(Analyte) %>%
  mutate(
    Q1  = quantile(Value, 0.25, na.rm = TRUE),
    Q3  = quantile(Value, 0.75, na.rm = TRUE),
    IQR = Q3 - Q1,
    is_outlier = Value < (Q1 - 1.5 * IQR) | Value > (Q3 + 1.5 * IQR)
  ) %>%
  ungroup()

ggplot(df_plt, aes(x = 1, y = Value)) +
  geom_jitter(
    aes(color = ifelse(is_outlier, 'Outlier', 'Normal')),
    shape = 21,
    width = 0.1,
    na.rm = TRUE
  ) +
  scale_color_manual(
    values = c('Normal' = 'lightgray', 'Outlier' = 'red')
  ) +
  geom_boxplot(
    outlier.shape = NA,
    na.rm = TRUE
  ) +
  facet_wrap(~Analyte, scales = 'free_y') +
  theme(legend.position = 'none')

```

There's a few, though none I would remove outright. However, further on I decided I wanted to remove a few extreme DissAmmonia, pH, Chlorophyll, and DissOrthophos values (7 total), which I'll do here:

```{r}
df_monthly_rm <- df_monthly_rm %>%
  filter(!(DissAmmonia > 0.4 | pH < 6.5 | Chlorophyll > 35 | DissOrthophos < 0.01) |
           is.na(DissAmmonia) |
           is.na(pH) |
           is.na(Chlorophyll) |
           is.na(DissOrthophos))
```

## Impute (Optional)

Next we'll impute missing data using a linear approximation. **If methods don't require a continuous, equally-spaced time series, this step is optional.** However, I like doing it (when feasible) to offset unbalanced design issues.

If your data has non-detects, make sure to uniformally impute those values between (slightly larger than) 0 and the reporting limit.

```{r}
df_monthly_imputed <- df_monthly_rm %>%
  arrange(Region, Date) %>%
  group_by(Region) %>%
  mutate(
    across(
      DissAmmonia:Total_Inflow,
      ~ is.na(.x),
      .names = 'imp_{.col}'
    ),
    across(
      DissAmmonia:Total_Inflow,
      ~ zoo::na.approx(.x, na.rm = FALSE),
      .names = '{.col}'
    )
  ) %>%
  ungroup()

values_long <- df_monthly_imputed %>%
  pivot_longer(
    cols = DissAmmonia:Total_Inflow,
    names_to = 'variable',
    values_to = 'value'
  )

flags_long <- df_monthly_imputed %>%
  pivot_longer(
    cols = starts_with('imp_'),
    names_to = 'flag_var',
    values_to = 'imp'
  ) %>%
  mutate(variable = sub('imp_', '', flag_var)) %>%
  select(Date, Region, variable, imp)

df_plot <- values_long %>%
  left_join(flags_long, by = c('Date', 'Region', 'variable'))

vars_imputed <- df_plot %>%
  group_by(variable) %>%
  summarise(any_imp = any(imp, na.rm = TRUE)) %>%
  filter(any_imp) %>%
  pull(variable)

regions <- unique(df_plot$Region)

for (r in regions) {
  p <- df_plot %>%
    filter(
      Region == r,
      variable %in% vars_imputed
    ) %>%
    ggplot(aes(Date, value)) +
      geom_line(color = 'gray40', linewidth = 0.5) +
      geom_point(aes(color = imp), size = 1.5, shape = 21, stroke = 0.8) +
      scale_color_manual(
        values = c('FALSE' = 'black', 'TRUE' = 'red'),
        labels = c('Observed', 'Imputed'),
        name = ''
      ) +
      facet_wrap(~ variable, scales = 'free', ncol = 2) +
      labs(
        title = paste('Region:', r)
      )

  print(p)
}

# clean dataframe
df_monthly_imputed <- df_monthly_imputed %>%
  select(-starts_with('imp_'))

```

## Transform

Next we transform the data. Here I do:

-   log transforms: helps stabilize variance, fits certain data patterns

-   AR1 lag terms: so previous time's value can be used in regressions. I just do this for response variables (Chlorophyll, log_Chlorophyll, and later log_DissAmmonia).

```{r}
# log data
df_monthly_log <- df_monthly_imputed %>%
  mutate(
    across(
      DissAmmonia:Total_Inflow,
      ~ log(.x),
      .names = 'log_{.col}'
    )
  )

# add lags
df_monthly_log <- df_monthly_log %>%
  arrange(Region, Date) %>%
  group_by(Region) %>%
  mutate(
    lag_logChlorophyll = lag(log_Chlorophyll),
    lag_Chlorophyll = lag(Chlorophyll),
    lag_logDissAmmonia = lag(log_DissAmmonia)
  ) %>%
  ungroup() %>%
  relocate(c(Chlorophyll, log_Chlorophyll, Date),
         .after = last_col())
```

# Model Construction

Now we can figure out what analytes we want in our models.

SEMs are essentially composed of multiple linked models. **For each one, it's important to carry out the next few steps (Variable Relationships and Test Models).**

Since Chla is my main response variable, I'll be sticking to that in these examples.

# Variable Relationships

## Correlations

First we look to see if any of our data is highly correlated (or not) since it can inform what we use in the model.

```{r}
corrplot::corrplot(
  cor(df_monthly_log %>% select(c(Chlorophyll,DissAmmonia:Total_Inflow)), use = 'pairwise.complete.obs'),
  method = 'color',
  type = 'lower',
  diag = TRUE
)
```

```{r}
corrplot::corrplot(
  cor(df_monthly_log %>% select(c(Chlorophyll,DissAmmonia:Total_Inflow)), use = 'pairwise.complete.obs'),
  method = 'number',
  number.cex = 0.6,
  type = 'lower',
  diag = TRUE
)
```

DissolvedOxygen and Temperature are strongly negatively correlated; DO is also correlated with DissNitrateNitrite. Temp is also slightly correlated with DissAmmonia and, more strongly, DissNitrateNitrite. DON/DissNitrateNitrite and Turbidity/Total Inflow are moderately correlated, which makes sense.

## Bivariate (Continuous)

Next we graph our explanatory variables against our response variable. I like to look at the raw points (to get a true sense of the data) and interpolated splines to understand the shape of relationship, including by region since that's a possible variable.

I ended up using log_Chlorophyll as my response, but kept the Chlorophyll graphs for reference.

```{r}
df_plt <- df_monthly_log %>%
  pivot_longer(
    cols = c(DissAmmonia:Total_Inflow, lag_Chlorophyll),
    names_to = 'Analyte',
    values_to = 'Value'
  )

df_plt_log <- df_monthly_log %>%
  pivot_longer(
    cols = c(log_DissAmmonia:log_Total_Inflow, lag_logChlorophyll),
    names_to = 'Analyte',
    values_to = 'Value'
  )
```

### Scatterplots

```{r}
ggplot(df_plt) +
  geom_point(aes(Value, Chlorophyll, color = Region), shape = 21, alpha = 0.5) +
  facet_wrap(~Analyte, scales = 'free_x')
```

```{r}
ggplot(df_plt_log) +
  geom_point(aes(Value, Chlorophyll, color = Region), shape = 21, alpha = 0.5) +
  facet_wrap(~Analyte, scales = 'free_x')
```

```{r}
ggplot(df_plt) +
  geom_point(aes(Value, log_Chlorophyll, color = Region), shape = 21, alpha = 0.5) +
  facet_wrap(~Analyte, scales = 'free_x')
```

```{r}
ggplot(df_plt_log) +
  geom_point(aes(Value, log_Chlorophyll, color = Region), shape = 21, alpha = 0.5) +
  facet_wrap(~Analyte, scales = 'free_x')
```

### Splines (Global)

```{r}
ggplot(df_plt) +
  geom_smooth(aes(Value, Chlorophyll)) +
  facet_wrap(~Analyte, scales = 'free_x')
```

```{r}
ggplot(df_plt_log) +
  geom_smooth(aes(Value, Chlorophyll)) +
  facet_wrap(~Analyte, scales = 'free_x')
```

```{r}
ggplot(df_plt) +
  geom_smooth(aes(Value, log_Chlorophyll)) +
  facet_wrap(~Analyte, scales = 'free_x')
```

```{r}
ggplot(df_plt_log) +
  geom_smooth(aes(Value, log_Chlorophyll)) +
  facet_wrap(~Analyte, scales = 'free_x')
```

### Splines (Regional)

```{r}
ggplot(df_plt) +
  geom_smooth(aes(Value, Chlorophyll, color = Region)) +
  facet_wrap(~Analyte, scales = 'free_x')
```

```{r}
ggplot(df_plt_log) +
  geom_smooth(aes(Value, Chlorophyll, color = Region)) +
  facet_wrap(~Analyte, scales = 'free_x')
```

```{r}
ggplot(df_plt) +
  geom_smooth(aes(Value, log_Chlorophyll, color = Region)) +
  facet_wrap(~Analyte, scales = 'free_x')
```

```{r}
ggplot(df_plt_log) +
  geom_smooth(aes(Value, log_Chlorophyll, color = Region)) +
  facet_wrap(~Analyte, scales = 'free_x')
```

From these, I decided to keep log_DissAmmonia, Total_Inflow, and Temperature in my model. I also added log_DissOrthophosphate since I think it's usually considered an important variable. Note that Temperature appears to have a cubic relationship with log_Chlorophyll.

## Bivariate (Categorical)

It's also worth seeing how chlorophyll is influenced by categorical variables, in this case region:

```{r}
ggplot(df_plt) +
  geom_boxplot(aes(Region, Chlorophyll, fill = Region))
```

```{r}
ggplot(df_plt) +
  geom_violin(aes(Region, Chlorophyll, fill = Region))
```

South Central has the most variance, followed by Suisun Marsh. The other regions seem similarly distributed.

# Test Models

Since SEMs are a series of related models (with shared covariance), it's worth it to check each sub-model individually. For traditional SEMs, only `lm` (fixed effects) models can be used, though piecewiseSEM can handle other model types. Here, I used linear mixed effects models.

For simplicity, I subset the data from 02-01-2017 to 12-31-2022 to escape issues with unbalanced regional data. I also added three new variables: centered Temperature and the squared/cubed versions of it. This is so I correctly model Temperature's likely non-linear dependence on Chlorophyll.

I also remove some extreme outliers I found in the model.

```{r}
df_monthly_subset <- df_monthly_log %>%
  filter(Year < 2023 & Date > '2017-02-01')

df_monthly_subset <- df_monthly_subset %>%
  slice(-c(5, 337))

df_monthly_subset <- df_monthly_subset %>%
  mutate(       
    Temperature_c = scale(Temperature, center = TRUE, scale = FALSE)[,1],
    Temperature2 = Temperature_c^2,
    Temperature3 = Temperature_c^3
    )

df_monthly_subset <- df_monthly_subset %>%
  drop_na(log_Chlorophyll, log_Salinity, log_DissAmmonia, log_DissOrthophos, log_Turbidity,
          lag_logChlorophyll, lag_logDissAmmonia, log_pH, log_Temperature)
```

## LMM - Chla

### Test vs Train split

*Usually not needed for causal inference, mostly just doing as a quick check that the model makes sense. Will re-run with full dataset.*

Temperature looked cubic to me, so I centered it and added polynomial terms. It's coefficients must be interpreted together to give you the slope of T, which is not a constant (B_1 + 2B_2(T_c) + 3B_3(T_c\^2)).

```{r}
# --- split testing/training ---
sorted_dates <- sort(unique(df_monthly_subset$Date))

cut_idx <- floor(0.8 * length(sorted_dates))

train_dates <- sorted_dates[1:cut_idx]
test_dates  <- sorted_dates[(cut_idx + 1):length(sorted_dates)]

df_train <- df_monthly_subset %>% filter(Date %in% train_dates)
df_test  <- df_monthly_subset %>% filter(Date %in% test_dates)

# -- training model --
fit_main <- lmer(
  log_Chlorophyll ~ 
    log_Salinity + 
    log_DissAmmonia +
    log_DissOrthophos +
    log_Turbidity +
    lag_logChlorophyll +
    pH +
    Temperature_c +
    Temperature2 +
    Temperature3 +
    (1|Region),
  data = df_train,
  na.action = na.omit
)

summary(fit_main)

# -- predicted model --
df_test <- df_test %>%
  drop_na(log_Chlorophyll, log_Temperature, log_DissAmmonia, log_Total_Inflow)

df_test <- df_test %>%
  mutate(pred_log_Chl = predict(fit_main, newdata = df_test))

# -- plot --
rng <- range(c(df_test$pred_log_Chl, df_test$log_Chlorophyll), na.rm = TRUE)

ggplot(df_test, aes(x = pred_log_Chl, y = log_Chlorophyll)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = 'lm', se = FALSE, linewidth = 0.7) +
  geom_abline(intercept = 0, slope = 1, linetype = 'dashed') +
  coord_equal(xlim = rng, ylim = rng) +
  labs(
    x = 'Predicted',
    y = 'Observed',
    title = 'Predicted vs Observed'
  ) +
  theme_minimal()

RMSE <- sqrt(mean((df_test$log_Chlorophyll - df_test$pred_log_Chl)^2))
MAE  <- mean(abs(df_test$log_Chlorophyll - df_test$pred_log_Chl))
bias <- mean(df_test$pred_log_Chl - df_test$log_Chlorophyll)

print(RMSE); print(MAE); print(bias)

plot(fit_main)
```

Looks pretty good! It does overestimate, but it's a consistent offset and not horribly off.

Next is model diagnostics:

```{r}
plot(acf(residuals(fit_main)))
plot(pacf(residuals(fit_main)))
```

```{r}
infl <- performance::check_outliers(fit_main)
infl
```

```{r}
res <- residuals(fit_main)

qqnorm(res)
qqline(res, col = 'red', lwd = 2)
```

Model diagnostics look pretty good; slight autocorrelation at the 15th lag that's indicates minor model misspecification, and there's some right-tail skew, but neither are particularly noteworthy.

### Full model

Now we run the model:

```{r}
fit_main <- lmer(
  log_Chlorophyll ~ 
    log_Salinity + 
    log_DissAmmonia +
    log_DissOrthophos +
    log_Turbidity +
    lag_logChlorophyll +
    log_pH +
    Temperature_c +
    Temperature2 +
    Temperature3 +
    (1|Region),
  data = df_monthly_subset,
  na.action = na.omit
)

summary(fit_main)
plot(fit_main)
```

The coefficients are similar, which is expected.

```{r}
plot(acf(residuals(fit_main)))
plot(pacf(residuals(fit_main)))
```

```{r}
infl <- performance::check_outliers(fit_main)
infl
```

```{r}
mf <- model.frame(fit_main) 
row_map <- as.numeric(rownames(mf))
res <- residuals(fit_main)

df_resrow <- tibble(
  row = row_map,
  resid = res,
  fitted = fitted(fit_main)
)

df_res <- df_monthly_subset %>%
  mutate(row = row_number()) %>%
  left_join(df_resrow, by = 'row')
```

```{r}
res <- residuals(fit_main)

qqnorm(res)
qqline(res, col = 'red', lwd = 2)
```

Looks similar to the test model, so we're good.

## LMM - DissAmmonia

Since SEMs ideally have at least two "sub models", I'll throw together a quick model for DissolvedAmmonia. I didn't bother with intense model diagnostics here, but they should definitely be done for real models.

```{r}
df_plt_log <- df_monthly_log %>%
  mutate(lag_logDissAmmonia = lag(log(DissAmmonia))) %>%
  pivot_longer(
    cols = c(log_DissNitrateNitrite:log_Total_Inflow, lag_logChlorophyll, lag_logDissAmmonia),
    names_to = 'Analyte',
    values_to = 'Value'
  )
```

```{r}
ggplot(df_plt) +
  geom_smooth(aes(Value, log_DissAmmonia)) +
  facet_wrap(~Analyte, scales = 'free_x') +
  theme_bw()
```

```{r}
ggplot(df_plt_log) +
  geom_smooth(aes(Value, log_DissAmmonia)) +
  facet_wrap(~Analyte, scales = 'free_x') +
  theme_bw()
```

```{r}
fit_amm <- lmer(
  log_DissAmmonia ~ 
    lag_logDissAmmonia +
    Temp1 +
    log_Turbidity +
    (1|Region),
  data = df_monthly_subset,
  na.action = na.omit
)

summary(fit_amm)
plot(fit_amm)
```

# SEMs

Linear piecewise model (direct comparison to tradtional SEM):

```{r}
library(piecewiseSEM)

fit_main <- lm(
  log_Chlorophyll ~ log_Salinity + 
    log_DissAmmonia +
    log_DissOrthophos +
    log_Turbidity +
    lag_logChlorophyll +
    Temperature_c +
    Temperature2 +
    Temperature3,
  data = df_monthly_subset)

fit_amm <- lm(
  log_DissAmmonia ~ lag_logDissAmmonia +
    Temperature_c +
    log_Salinity +
    log_Turbidity +
    log_DissOrthophos,
    data = df_monthly_subset)

psem_out <- psem(fit_main, fit_amm, data = df_monthly_subset)
summary(psem_out, standardize = 'scale')
```

LMM piecewise model:

```{r}
fit_main <- lmer(
  log_Chlorophyll ~ log_Salinity +
    log_DissAmmonia +
    log_DissOrthophos +
    log_Turbidity +
    lag_logChlorophyll +
    Temperature_c +
    Temperature2 +
    Temperature3 +
    (1|Region),
  data = df_monthly_subset)

fit_amm <- lmer(
  log_DissAmmonia ~ lag_logDissAmmonia +
    Temperature_c +
    log_Turbidity +
    (1|Region),
  data = df_monthly_subset)

psem_out <- psem(fit_main, fit_amm, data = df_monthly_subset)
summary(psem_out, standardize = 'scale')
```

Linear model (traditional SEM):

```{r}
library(lavaan)

model <- '
  # regression paths
    log_Chlorophyll ~ log_Salinity + 
    log_DissAmmonia +
    log_DissOrthophos +
    log_Turbidity +
    lag_logChlorophyll +
    Temperature_c +
    Temperature2 +
    Temperature3
    
    log_DissAmmonia ~ lag_logDissAmmonia +
    Temperature_c +
    log_DissOrthophos +
    log_Turbidity

    log_DissAmmonia ~~ log_Chlorophyll
'

fit <- sem(model, data = df_monthly_subset)

summary(fit, fit.measures = TRUE, standardized = TRUE)
```

```{r}
library(semPlot)

semPaths(
  fit,
  what = 'std',      
  whatLabels = 'std',
  layout = 'tree2', 
  nCharNodes = 5,
  edge.label.cex = 1.1,
  sizeMan = 5,         
  edge.width = 1,
  esize = 1,
  residuals = FALSE,   
  mar = c(7, 7, 7, 7)
)
```
